In vitro sexual dimorphism establishment in schistosomes

Version 1

Abstract

Schistosomes are parasitic flatworms that cause Schistosomiasis, a major Neglected Tropical Disease (NTD) that affects more than 250 million people worldwide. With two morphologically distinct sexes, a heterogametic female (ZW) and a homogametic male (ZZ), schistosomes are an exception among flatworms, which are largely hermaphroditic. Sexual dimorphism in schistosomes only becomes apparent by adulthood within the mammalian host. The cellular and molecular mechanisms underlying the sexual differentiation of these parasites are poorly understood, partly due to intrinsic challenges in assessing parasite development in vivo. Therefore, robust and reproducible approaches for maintaining and developing parasites in vitro may help to overcome these difficulties. However, to date, only a few studies have focused on protocols that allow cultured parasites to reach sexual dimorphic stages, and none have been reproduced, limiting the ability to understand the unique sexual biology of this major human parasite. Here, we refine a protocol for long-term culture of newly transformed cercariae that developed in vitro into sexually dimorphic forms. We assessed the effect of two different sera, Foetal Bovine Serum (FBS) and Human Serum (HS), added to the culture medium supplemented with human red blood cells. We found that in contrast to FBS-cultured worms, HS-cultured parasites digested red blood cells, a crucial step for long term parasite culture. Additionally, while most FBS-cultured parasites did not progress beyond an early liver stage, sexual dimorphism was clearly established in the HS-cultured worms, albeit delayed compared to in vivo development. Moreover, EdU pulse-chase experiments revealed a continuous proliferation of parasite cells over time in HS-cultured parasites, while a significantly lower number of proliferating cells were detected in FBS-cultured worms. This protocol paves the way to study parasite development in vitro, positively impacting the principles of the 3Rs (Replacement, Reduction and Refinement) for animal research, and allowing for in-depth studies of sexual dimorphism establishment as well as in vitro screening for novel control strategies across the life cycle of these major human parasites.

Introduction

Schistosomiasis, a major Neglected Tropical Disease (NTD) affecting >250 million people worldwide, is caused by the infection with blood flukes (class Trematoda) in the genus Schistosoma1. The pathology associated with schistosomiasis is largely driven by egg trapping in different tissues depending on the species. The infection with Schistosoma mansoni leads to eggs lodged mainly in the liver inducing inflation, fibrosis and granuloma formation2. This suggests that interfering with the sexual development of these parasites could potentially restrict human pathology. Currently, the complete reliance on a single drug (Praziquantel), to treat the infection and its use in drug mass administration programs in endemic areas threatens the development of drug resistance3. Therefore, identifying novel targets for alternative, more efficient control strategies is a priority that could be addressed by elucidating mechanisms involved in schistosome development, including the establishment of sexual dimorphism.

While most of the parasitic flatworms are hermaphrodites, schistosomes are dioecious with genetically determined female (2n=16, ZW) and male (2n=16, ZZ) individuals. Whilst male and female parasites are morphologically indistinguishable across most developmental stages, sexual dimorphism becomes apparent by adulthood within the mammalian host4. Male and female worms undergo separate but concurrent sexual differentiation of their gonads and somatic tissues that eventually allows intersexual pairing, a critical step for female maturation, egg production and life cycle propagation5,6. Transcriptomic studies, at both the bulk7–11 and single cell12–14 levels, have revealed key pathways and molecular mediators underlying this critical process of female sexual maturation induced by pairing. However, the cellular and molecular mechanisms driving sexual dimorphism prior to pairing remain poorly characterised, partly due to intrinsic challenges in assessing the development of the parasite in vivo11.

Approaches for better understand the biology of parasites at large have been developed and optimised using in vitro or ex vivo refined culture systems15,16. Combinations of culture media components, sera from different sources and additives have been tested17,18. In addition, organ-on-a-chip systems and organoid-based 2D and 3D culture platforms19,20 have first been implemented for protozoa parasites such as Plasmodium, Toxoplasma, and Cryptosporidium species. These methods have facilitated the molecular dissection of processes underlying single-cell parasite development and interaction with the host21–23. Recently, these technologies have started to be transferred to metazoan parasites to study host-parasite interaction and immune response in the context of helminth infections24–28. In vitro cultivation of both larval tapeworms (cestodes) and stem cells isolated from cestodes, coupled with ‘omic’ and imaging analyses have unveiled the role of stem cells in parasite development29,30. However, despite these significant developments in the field, reproducing the natural development of trematodes under controlled in vitro conditions remains extremely challenging16.

Significant progress was achieved in culturing schistosomes when, more than forty years ago, Paul F. Basch developed and optimised comprehensive culture protocols for intra-mammalian developmental stages of schistosomes31–33. Basch reported for the first time the development of male and female parasites from cercariae, with worm pairing and production of infertile eggs entirely in vitro 31. Remarkably, to the best of our knowledge, these results have not been replicated and no further attempt to obtain egg-laying worm pairs developed in vitro from cercariae has been reported since Basch’s seminal studies16. More recently, in vitro culture protocols for schistosomes have been refined either to maintain ex vivo parasites collected from infected mice to study parasite reproduction biology34,35, or to culture juvenile parasites from cercariae and develop drug screening approaches36–39. However, no long-term culture conditions to specifically assess schistosome sexual differentiation have been reported since 198132. Despite one research group referring to the approach40, there has been no detailed description and no widespread adoption. In fact, published culturing studies have continued to utilise FBS34,41.

Novel functional approaches applied to culture systems that allow reliable and reproducible in vitro establishment of sexual dimorphism will shine new light into molecular mechanisms underlying schistosome intra-mammalian development. Here, we refined a protocol for long-term culture of newly transformed cercariae by assessing the effect of two different sera, Foetal Bovine Serum (FBS) and Human Serum (HS), added to the medium supplemented with human Red Blood Cells (hRBCs). Striking differences were evident between parasites maintained in either FBS or HS. First, hRBC were digested and hemozoin became apparent in the intestines of HS-cultured parasites within 5 days in contrast to FBS-cultured parasites that mostly lacked visible hemozoin. Second, while most of the FBS-cultured parasites did not progress beyond lung and early liver stage, HS-cultured parasites reached sexually dimorphic stages, albeit at a slightly delayed rate compared to in vivo development. Furthermore, phenotypic differences between FBS- and HS-cultured parasites became evident as early as 48 hours in culture, with HS-cultured parasites exhibiting higher rates of cell proliferation. Taken together, this protocol allows early sexual development of schistosomes to be studied in vitro, provides a method for high throughput drug screening targeting parasite development, and reduces the reliance on mammalian hosts in research.

Results

Sexually dimorphic schistosomes developed entirely in vitro from cercariae

Searching for culture media able to cultivate parasites from cercariae to maturity, we decided to test Human Serum (HS) compared to the commonly used Foetal Bovine Serum (FBS). This rationale is based both on the fact that humans are the definitive host of schistosomes, and that two previous reports in 1981 showed that HS was capable of producing mature schistosomes in vitro31,32, but they were never replicated. Therefore, to ascertain the in vitro sexual dimorphism establishment of schistosomes entirely developed from cercariae, we compared the effect of two different sources of blood serum: FBS and HS. The development of schistosomula derived from mechanically transformed cercariae was assessed in at least 15 independent experiments, five of which were maintained over a period of at least 10 weeks (Figure 1A; Supplementary Table S1).

Dimorphic female and male schistosomes entirely developed in vitro from cercariae.

A. Schematic representation of the collection and mechanical transformation of cercariae into schistosomula for long-term in vitro culture. B. Morphological scoring of cultured S. mansoni schistosomula at the indicated time points after in vitro transformation (week 1 to week 10) for worms in Long Term Culture (LTC) medium supplemented either with Human Serum (HS - left) or Foetal Bovine Serum (FBS - right). Heatmap columns represent five distinct morphological categories, and rows indicate 5 independent culture experiments, i.e. parasites obtained from different batches of infected snails. Heatmap colours represent the percentage of worms for each replicate in each morphological category. Middle panel: Representative images of in vitro schistosomula cultured in either HS or FBS as indicated. Scale bars: 100 µm. Category 1: Early schistosomula; Category 2: Lung schistosomula; Category 3: Early liver schistosomula; Category 4: Late liver schistosomula; Category 5: Dimorphic schistosomula.

No morphological differences were observed between parasites cultured either in FBS or HS within the first week in culture; in both conditions most parasites (76% in FBS and 72% in HS in average) were classified as early schistosomula (Supplementary Figure S1) with few lung and early liver schistosomula (Figure 1B, Week 1). The mean mortality at Week 1 was slightly higher, but not statistically significant (P= 0.8), in worms cultured in HS (11.3 ± 2.7%) compared to the mortality registered in FBS-cultured parasites (5 ± 2.3%), consistent with previous findings38.

From Week 2 onwards, differences in parasite development between the two conditions became apparent (Figure 1B). Parasites cultured in FBS were no longer able to develop beyond early liver stage (14.8% in average), with the vast majority of worms (79% in average) still stunted and classified as early and lung schistosomula. The mortality rate of FBS-cultured parasites continued to increase, reaching an average of ∼80% by Week 10, after which the experiments under this condition were stopped as most parasites were dead (Supplementary Figure S2). On the other hand, parasites cultured in HS further developed over time throughout all categories, achieving marked sexual dimorphism by Week 6 (Figure 1B; Supplementary Figure S3A), confirmed by PCR (Supplementary Figure S3B; Supplementary Table S2). As previously described for the in vivo development of schistosomes11, the in vitro cultured parasite showed developmental asynchrony in agreement with Basch’s observations32; however, by Week 10 the majority of the worms (73.7 ± 25.4%) have acquired sexual dimorphism (Figure 1B).

The differences in parasite development between the two tested conditions were also confirmed by measuring worm areas at different time points in culture (Figure 2; Supplementary Table S3). Parasites cultured in the presence of HS grew exponentially over time to reach ∼0.1 mm2, ∼24-fold larger than parasites cultured in FBS for 10 weeks, which only grew slightly, plateauing at an average 0.005 mm2, similar to the size reached by parasites after only two weeks in HS (Figures 1 and 2).

Parasites cultured in Human Serum grew in size unlike those cultured in Foetal Bovine Serum.

Bar plot representing area measurements of schistosomula developed in vitro and cultured in media complemented either in Foetal Bovine Serum (FBS) or Human Serum (HS) at indicated weeks after cercarial transformation. Bars indicate mean area (µm²) ± SEM. HS: Human Serum (light brown); FBS: Foetal Bovine Serum (blue). *: P-value < 0.05 in indicated pairwise comparisons.

Parasites developed in human serum readily digest red blood cells

It is widely accepted that within the mammalian host schistosomes begin to feed on blood cells through their digestive tract at 10 days post-infection; at this time the majority of the parasites have left the lungs and started to colonise the portal system veins in the liver42. On the 13th day of culture, human Red Blood Cells (hRBCs) were added to the parasites to allow them to feed and thus continue their development. At this point, they begin to capture and degrade erythrocytes producing hemozoin, a black pigment derived from host haemoglobin degradation and visible in the worms’ intestines. Even though few parasites in FBS reached the early liver stage within the first week in culture, a minority of them were able to digest hRBCs (3.6 ± 4.7 %), indicated by displaying black guts (BG+, Figure 3). In contrast, a significant proportion of HS-cultured parasites (36.2 ± 33.6 %) had already reached the early liver stage by the second week in 3 out of 5 experiments (Figure 1B, Week 2). Moreover, parasites cultured in HS displayed a functional digestive system capable of assimilating hRBCs; more than half of the HS-cultured parasites (65 ± 6 %, P-value < 0.05) were BG+ in comparison to FBS-cultured parasites (3.7 ± 4.7 %) (Figure 3; Supplementary Table S4).

Parasites developed in Human Serum readily digest red blood cells.

A. Bar Plot representing the percentage of Human Serum (HS)- or Foetal Bovine Serum (FBS)-cultured schistosomula with (BG+, light brown) or without (BG-, blue) black guts due to the presence of intestinal hemozoin. Washed human Red Blood Cells (hRBCs) were added into the media at day 13 post-transformation and images captured one or two days later. Error Bars = SEM. B. Representative images of in vitro developed schistosomula cultured in FBS or HS one day after adding hRBC (+ RBC) and controls without RBC (- RBC). Scale bars: 100 µm.

Development of parasites in human serum may be driven by stem cell proliferation

The growth and development of an organism is driven by a finely regulated combination of cell hyperplasia, hypertrophy, and apoptosis across different tissues43. In schistosomes, a complex stem cell system comprised of both somatic and germline stem cells has been described by leveraging recent single cell transcriptomics44. Therefore, we decided to investigate whether the differences in development, growth and feeding capacity between parasites cultured in either FBS- or HS-complemented medium were associated with distinct cellular proliferation rates. EdU pulse-chase experiments revealed notably lower cell proliferation in FBS-cultured parasites as early as 2 days post transformation (Figure 4). It has previously been demonstrated that a group of 5 well-defined somatic stem cells is the only set of cells that actively proliferate in 2-day schistosomula45. Hence, the proliferating cells observed and quantified in our study were most likely stem cells (Supplementary Figure S4; Supplementary Video S1). The difference between the number of proliferating stem cells in parasites cultured in FBS or HS increased significantly over time. HS-cultured schistosomula showed higher numbers of proliferating stem cells, with an average of >60 EdU+ cells per worm (Figure 4), whereas most FBS-cultured parasites remained stunted in the lung or early liver stages. These stunted worms displayed no more than an average of 20 EdU+ cells per worm (Figure 4). hRBCs were added at day 13 post-cercarial transformation to both FBS- and HS-complemented culture media. Worms kept in FBS or HS in the absence of hRBCs were included as controls. Regardless of the serum employed, no significant differences in the numbers of proliferating cells were observed between worms cultured in the presence or absence of hRBCs (Supplementary Tables S5 and S6).

Development of parasites in human serum may be driven by stem cell proliferation.

A. Violin plots showing the number of Edu+ cells per worm at indicated time points (2, 8, and 15 days post cercarial transformation) in parasites cultured either in Foetal Bovine Serum (FBS, blue) or Human Serum (HS, light brown). Human Red Blood Cells (hRBCs) were added in the culture at day 13 post cercarial transformation. The small black dots indicate individual worms, and the big black point indicates the mean of EdU+ cells per worm. Statistical analysis was performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 (*) considered significant. B. Representative images of parasites displaying Edu+ cells at each indicated time point and culture condition. Edu+ cells and nuclei were labelled with Alexa fluor 488 (green) and DAPI (white/grey), respectively. Scale bars: 50 µm or 75 µm as indicated.

In vitro cultured schistosomes display sexual dimorphism, developing reproductive systems and pairing capacity

In vitro-developed schistosomes began to show sexually dimorphic features from day ∼42 onwards in HS-complemented culture medium (Figure 1, Week 6). Confocal microscopy of DAPI- and Phalloidin-stained in vitro cultured male worms confirmed the presence of gynaecophoric canal, developing 3 to 5 testis lobes containing sperm cells, and cirrus (Figure 5A-D; Supplementary Video S2). Likewise, we confirmed the presence of primordial ovaries, oviduct, ootype and uterus in female parasites entirely developed in vitro (Figure 5E-H; Supplementary Video S3). In some experiments sexually dimorphic parasites cultured in HS medium were kept alive for more than 150 days (Supplementary Figure S5). While the establishment of sexual dimorphism was robust and reproducible across more than 15 independent experiments, pairing between male and female parasites was rare. Pairing was observed in ∼7% of the experiments, and usually after day ∼80 in culture (Figures 6A and 6B; Supplementary Video S4).

In vitro cultured schistosomes display sexual dimorphism and developing reproductive systems.

Representative confocal microscopy images of in vitro developed male (A-D) and female (E-H) at day 60 in culture. Worms shown in panels G and H are different individuals. CellMask Green Actin Tracking Stain: green, DAPI: grey (A-E, G), cyan (F, H). C: Cirrus. T: Testis, OS: Oral sucker. VS: Ventral sucker. GC: Gynaecophoric canal. G: Gut. GP: Genital pore. U: Uterus. O: Ovary. Oo: Ootype. OV: Oviduc. S: sperm. Scale bar: 50 µm.

In vitro cultured schistosomes are capable of pairing.

A, B: Representative bright-field images of pairs of schistosomes developed entirely in vitro after 80 (A) and 150 (B) days of culture in culture medium supplemented with HS. Scale bars: 500 µm (A), 100 µm (B). C, D: Representative bright-field images of a worm pairs between ex vivo-collected males and in vitro-developed females (C) and ex vivo-collected females and in vitro-developed males (D) within 24 hours after placing the worms in the same well to facilitate pairing. Scale bars: 100 µm, E, F: Confocal microscopy image of a schistosome pair (ex vivo-collected male and in vitro-developed female) in copula (E), and magnification of the ovarian area, highlighting maturing oocytes (F). CellMask Green Actin Tracking Stain: green, DAPI: grey, CellMask Deep Red Plasma Membrane Stain: magenta. Scale bar: 150 µm (E), 50 µm (F).

Considering the rarity of the pairing between in vitro developed worms, we investigated whether these parasites display the capacity of pairing with in vivo developed worms collected from experimentally infected mice. Male and female adult worms developed in vivo and recovered from mice by portal perfusion on day 42 post-infection, were sorted by sex and placed in the same tissue culture plate well with worms of the opposite sex developed in vitro (>70 days). Within 24 hours after co-culturing the in vitro developed with ex vivo collected worms, couples were observed (Figure 6C, D; Supplementary Video S5). These findings suggested that the in vitro developed and sexually dimorphic parasites are capable of intersexual pairing. Moreover, in vitro developed females coupled with ex vivo collected mature males displayed signs of primordial ovaries maturation with larger oocytes towards the posterior region of the ovary (Figure 6E, F; Supplementary Video S6). Remarkably, in more than 30 independent in vitro culture experiments, where male and female parasites developed sexual dimorphism and eventually paired up, no eggs were produced or laid. This indicates that further refinements in the culture protocol are needed to accelerate the parasite development (the in vitro development is delayed compared to the in vivo development), increase the likelihood of pairing, and facilitate the production of eggs.

Discussion

Recent progress in culture systems and functional genomic tools is allowing researchers to address long standing questions in helminth biology that were previously inaccessible to experimentation46–50. Among these enduring questions, stand those related to the unique sexual biology of the trematode family Schistosomatidae. Schistosomes are dioecious with genetically determined female and male individuals, which is unusual for flatworms4,51. However, the sexual dimorphism of male and female schistosome worms only becomes established within the mammalian host, and represents a critical step towards worm maturation, intersexual pairing, and egg production52. A better understanding of the molecular and cellular basis underlying sexual dimorphism establishment in schistosomes would lead to approaches to block parasite development and life cycle propagation. Studying the sexual development of schistosomes by performing controlled experiments in which parasites can be co-cultured, genetically manipulated or treated with different compounds requires robust and reproducible protocols for long-term in vitro culture.

Culture protocols have recently been developed to maintain ex vivo schistosomes collected from infected mice34,35, or to obtain juvenile parasites from cercariae for drug screening experiments36–39. However, no long-term culture conditions have been successfully refined to experimentally assess sexual differentiation and dimorphism establishment in schistosomes. Here, aiming to refine a culture medium formulation that supports in vitro parasite development and establishment of sexual dimorphism, we decided to compared the effect of modified Basch’s media18, supplemented with human Red Blood Cells (hRBCs) and 20% of either Fetal Bovine Serum (FBS) or Human Serum (HS). While initial parasite survival and development appeared comparable in both conditions during the first week in culture, striking morphological differences emerged from Week 2 onwards. Parasites cultured in HS, progressed through all developmental categories, and acquired sexual dimorphism by Week 6. On the other hand, parasites maintained in FBS were stunted at early stages (mainly lung stage). These experimental outcomes were consistent with the findings reported by Paul F. Basch31 in 1980. Probably, from the 1980s onwards a combination of factors that include ethical concerns, variability associated with human-derived products and safety considerations related to blood borne viruses such as HIV, and Hepatitis B and C, determined that parasitologists favoured the use of FBS over HS for helminth culture. FBS has been extensively used in cell culture media, refined and adapted to diverse human and animal cell types since 195853 providing a reliable source of amino acids, carbohydrates, hormones, lipids, proteins, vitamins, and growth factors54. Nevertheless, the scientific community is increasingly advocating for the replacement of FBS as a supplement in tissue culture55. This trend is driven by limitations associated with FBS, including the presence of undefined tentatively harmful factors for the cells in culture, lack of reproducibility, lack of transparency in its production, and critically ethical concerns related to animal welfare56.

Our protocol, that supports a fully in vitro development of schistosomes, would also positively impact the 3Rs (i.e., Reduction, Replacement, Refinement) for minimising the use of animals for research and serum production57,58.

Parasites cultured in the presence of HS not only developed into sexually dimorphic male and female worms, but strikingly, were able to digest hRBC and process haemoglobin within one day after the addition of hRBC. These findings may reflect differences in the gastrodermis development; worms cultured in the presence of HS may display a differentiated gastrodermis that enables haemoglobin digestion and accumulation of hemozoin within the intestines59. On the other hand, most parasites cultured in FBS were unable to produce hemozoin in the presence of hRBCs. This may be due to impaired gastrodermis differentiation that ultimately would lead to stunted development and death. Hematophagous parasites including Plasmodium species and schistosomes, obtain key nutrients via the proteolysis of host hemoglobin. However, this process leads to the production of free-heme groups which in turn generate highly toxic oxygen free radicals and lipid peroxidation60. These toxic free-heme derivatives become inactive when aggregated into an inert crystalline polymer, named hemozoin, observed as dark pigment within the intestines of schistosome intra-mammalian stages59. In addition, increasing evidence shows that hemozoin may play critical roles during parasite development by supplying iron for egg production59, and interaction with the mammalian host by an immune modulatory role61. In the mouse model, schistosomula begin to feed on blood once they have left the lungs and reached the portal system 9 to 11 days post infection42,62. Moreover, schistosomula collected from experimentally infected mice at day 13 post infection already show hemozoin in their guts11. Similarly, HS-developed schistosomula were able to feed on blood in vitro within a comparable time window; hRBCs were added in the medium on day 13 in culture, and within 24 hours most of the parasites’ gut contained hemozoin. These findings suggest that our in vitro culture system successfully recapitulates in vivo gastrodermis development of schistosomula. A better understanding of the in vitro development of the parasite gastrodermis and its role in the production of hemozoin will expose tentative novel targets for control63.

Although previous reports have shown that it is possible to cultivate parasites in cell-free environments38,64, adding hRBC has proven essential for long-term maintenance of parasites and for establishing sexual dimorphism. EdU pulse-chase experiments suggested that schistosome in vitro development in the presence of HS may be driven by the proliferation and differentiation of stem cells. The stem cell system in schistosomes has been extensively studied since the first description of somatic stem cells in adult worms65, followed by single cell transcriptomic identification and functional characterisation of three key stem cell populations in intra-snail stages45. Two of these stem cell types, maintained in intra-mammalian stages, proliferate and differentiate into precursors of somatic and germ line cells throughout development13. Although we have not performed transcriptomic analyses in this study, the positive correlation between number of EdU+ cells, and growth (determined by increasing worms’ area) and gross development in HS-developed parasites was evident already within the first 48 hours in culture. These findings indicate that stem cells may play a central role in driving organised growth, tissue differentiation and ultimately the establishment of the sexual dimorphism in HS-developed parasites. Conversely, FBS-cultured parasites displayed no more than 20 EdU+ cells per worm in average, reaching a plateau in the number of proliferating cells from day 8 in culture, which probably underlie the arrested development of these worms. Overall, the FBS-cultured parasites did not progress beyond the lung stage. Consistent with these findings, confocal microscopy imaging revealed in HS-developed parasites clear somatic and reproductive anatomical structures, including male gynaecophoric canals and testes, as well as female ovary primordia, oviducts, ootypes and uteri. These developmental features illustrate the capacity of our culture system to capture key biological transitions previously only accessible from experimentally infected animal models. That said, while our system was highly efficient in producing sexually dimorphic worms, spontaneous pairing between male and female parasites was extremely rare, mainly in aged in vitro cultures (from 80 to 100 days in culture) indicating that other factors, e.g., cholesterol, may be missing34. In any case, we decided to test the pairing capacity of the dimorphic worms by conducting pairing experiments between in vitro-developed females and ex vivo-collected males, and vice versa. Strikingly, couples were observed within 24 hours after co-cultivating in vivo and ex vivo opposite sex parasites, albeit no eggs were produced. However, we observed evidence of pairing-trigger oocyte maturation in in vitro-developed females paired with ex vivo-collected male worms. Our research group has recently been involved in the discovery and functional characterisation of a transcription factor of the retinoic acid receptor family, SmRAR, and related genes66. SmRAR and associated genes may play a key role in oocyte differentiation triggered by pairing. Our pairing experiments and confocal imaging analyses suggested that in vitro developed female oocytes started to differentiate after pairing with an in vivo developed male worm, probably mediated the activation of the SmRAR pathway66. Moreover, the involvement of a male-derived nonribosomal peptide pheromone, i.e., β-alanyl-tryptamine or ‘BATT’, in the pairing-driven female sexual maturation was demonstrated5. In addition, soluble factors produced by the worms with effect on the opposite sex cannot be ruled out6. It has been suggested that host-derived factors, such as the cytokine transforming growth factor β (TGF-β) may be critical for female reproductive development and embryogenesis67. More recently, other host-derived factors that include cholesterol and ascorbic acid have been shown to be critical for maintaining ex vivo fully mature females collected from infected mice34. Further research is needed to assess the role of these host- and parasite-derived cues in schistosome development.

In summary, we have demonstrated that the presence of HS is essential to support fully in vitro development of S. mansoni parasites from mechanically transformed cercariae to sexually dimorphic adults. Differences in the numbers of proliferating stem cells between worms developed in HS versus FBS were observed as early as 48 hours in culture. These findings may raise some concerns about studies that rely on in vitro culture protocols using FBS, including those focused on parasite developmental biology, interaction with the host, and drug screening. Human serum contains essential host-specific molecules that are absent or insufficient in FBS, and future efforts should aim to identify these factors. Optimising schistosome long-term culture systems will: (1) deepen our understanding of fundamental aspects of schistosome biology, development and host-parasite molecular crosstalk; (2) positively impact the 3Rs principles for animal research; and (3) reveal tentative targets for novel control strategies.

Material and Methods

Ethics statement

The complete life cycle of Schistosoma mansoni (NMRI strain) was maintained at the Wellcome Sanger Institute (WSI) and Aberystwyth University (AU), by infecting Biomphalaria glabrata snails and outbred mice (TO strain). All the animal regulated procedures were conducted under Home Office Project Licences P77E8A062 (WSI) and PP2955700 (AU), following the ARRIVE guidelines (https://arriveguidelines.org) and in accordance with guidelines and regulations stated by the UK Animals (Scientific Procedures) Act 1986 Amendment Regulations 2012. All the protocols were presented and approved by the Animal Welfare and Ethical Review Bodies (AWERB) of the WSI and AU. The AWERB is constituted as required by the UK Animals (Scientific Procedures) Act 1986 Amendment Regulations 2012. Human blood products, including serum and red blood cells were obtained from NHS Blood and Transplant (NHSBT), Non-Clinical Issue (NCI) services for research purposes. NHSBT-NCI provides donated material surplus to clinical requirements or unsuitable for therapeutic use that has been appropriately consented. The supply chain complies with all statutory and regulatory obligations including (but not limited to) the Human Tissue Act (2004) and associated Codes of Practice. The NHSBT-NCI products were acquired under the Study ‘In vitro development of the human parasite Schistosoma’, Integrated Research Application System (IRAS) project ID 319400, protocol number AU/DLS/011, Research Ethics Committee (REC) reference number 23/SS/0017.

Cercariae transformation and culture of schistosomula

Schistosoma mansoni (NMRI strain) schistosomula were obtained and cultured as previously described with minor modifications18. In brief, patent mix strain of Biomphalaria glabrata snails68 experimentally infected en masse with ∼20 miracidia per snail, were thoroughly rinsed, transferred to Lepple water (∼100 ml) and exposed to light for 2 h at 26 °C to induce cercarial shedding. Cercariae were collected, passed through a 100 μm filter into 50 ml tubes to remove snail debris and faeces, placed 1 h on ice, and concentrated by centrifugation (300 g for 3 min at 4°C with half break deceleration). Cercariae were washed 3 times in 1X PBS supplemented with 200 U/ml penicillin, 200 μg/ml streptomycin, 500 ng/ml amphotericin B (Merck). During the washes, the cercarial pellets were successively combined and finally resuspended in ‘Schistosomula wash medium’ (Dulbecco’s modified Eagle medium - DMEM, supplemented with 10 mM Hepes, 200 U/ml penicillin, 200 g/ml streptomycin, 500 ng/ml amphotericin B). Cercariae were vortexed full speed for 30 sec and placed on ice for 1 min. Thereafter, the cercarial tails were sheared off by ∼10 back and forth passes through a 19-gauge (19-G) needle, a ∼3 μl aliquot of parasites inspected under microscope to confirm that >90% of cercariae had their tail removed, and the cercarial heads were separated from the tails by a Percoll (Merck) gradient (1.5:1 percoll:DMEM) and centrifugation (300g for 15 min at 4 °C, half break deceleration). The pellet containing purified cercarial heads, was collected and washed 3 times by centrifugation (300g for 3 min at 4°C) in ‘Schistosomula wash medium’. Newly transformed schistosomula were counted in 12 5μl-aliquots under the microscope. To reduce parasite mortality due to a high density of worms, ∼8,000 schistosomula were transferred to each well of a 6-well tissue culture plate (Fisher Scientific, Loughborough, UK) containing 5 ml of modified Basch’s medium18 supplemented with additives and FBS or HS heat-inactivated FBS or HS serum, i.e., ‘Long-term culture medium’ (LTC) as indicated in Table 1. Parasites were cultured in a tissue culture incubator at 37°C under 5% CO2 in air. Washed human red blood cells (hRBCs) were added into the culture medium at indicated time points. LTC medium was replaced twice a week and washed hRBCs added to a final concentration of 0.02% v/v at indicated days after transformation. All human blood products, obtained from NHSBT-NCI, were handled with universal precautions within a biosafety cabinet class II and suitable PPE. The HS was heat-inactivated at 56°C for 30 min in an incubator, aliquoted and stored at −20°C until use. The hRBC were washed in ‘Schistosomula wash medium’ and stored at 4°C until use as previously described18. Briefly, the hRBCs were transferred from the pack to 50 ml tubes and washed by centrifugation at 500g for 5 min at 4°C. Half the supernatant was removed, tubes’ content combined, centrifuged as above and resuspended in ∼25 ml of ‘Schistosomula wash medium’. The washes were repeated 4 more times.

Phenotype scoring of developing schistosomula

To analyse the in vitro development of schistosomula, images and videos were taken at regular intervals of at least once a week in >10 independent culture experiments, using a digital Euromex camera connected to an Olympus CK2 optical microscope. Up to 18 pictures per experiment with an average of ∼400 worms per picture were used for parasite staging from at least 5 independent experiments (Supplementary Table S1). The investigators who assigned categories were blinded to experiment conditions. Based on a well-defined staging system11,32, the parasites were classified into 6 developmental categories as described38 (Supplementary Figure S1); category 0: dead parasites showing granulation, rounded shape and degraded or broken tegument - the number of dead parasites may be underestimated due to the loss of some dead worms during the media change; category 1: newly transformed schistosomula, corresponding to the in vivo ‘skin stage’, these worms retain the cercarial head shape with no visible internal structures such as gut; category 2: lung stage schistosomula, elongated and slim worms, usually with a bulged end; category 3: early liver stage, bigger and wider worms compared to the previous stage, with hemozoin pigment (i.e., haemoglobin degradation product) and two ceca gut; category 4: late liver stage, in which the gut is further developed, the two intestinal ceca have fused behind the ventral sucker, and the parasites started to acquire an evident vermiform shape; category 5: sexually dimorphic stage, comprising large, vermiform male- and female-looking worms. The females are longer and thinner than males, the ceca fusion localises closer to the anterior end of the worm (i.e., ∼⅓ of the worm length from the anterior end), and both suckers are smaller compared to the male suckers. The male worms are characterised by a larger and wider head, larger suckers, the ceca fusion closer to the posterior end of the animal (i.e., ∼⅔ of the worm length from the anterior end), and a clearly visible gynecophoral groove11,32.

To quantify positive or negative parasites for the presence of hemozoin within their intestines, i.e., black guts + (BG+) or BG- parasites, respectively, hRBCs were added to the culture at day 13, and one or two days later bright field microscopy images were taken. Counts of BG+ and BG- parasites were collected from 5 independent in vitro culture experiments. Statistical analysis was performed by ANOVA followed by Tuckey’s post-hoc multiple pairwise comparison with P≤0.05 considered significant (Supplementary Table S6).

Measurement of parasite area

Images of in vitro developing parasites were taken at weeks 1, 2, 3, 4, 6, 8 and 10 after cercariae transformation with an Euromex camera fitted to an Olympus CK2 optical microscope at 4x and 10x magnification. For these magnifications images of scale bars were captured for pixel to area values (µm²). Parasites from 5 independent long-term culture experiments were masked in COCO format from captured images using the CVAT (Computer Vision Annotation Tool) online server (https://www.cvat.ai/). Area metrics were then calculated for each parasite by leveraging the pycocotools package in a customised Python script. Raw and processed data is provided in Supplementary Table S3. Normal distribution was checked using Shapiro test and Mann Whitney tests performed with a significant value set up for P≤0.05 (Supplementary Table S6).

PCR for sexing and confirmation of sexual dimorphism

To confirm the in vitro establishment of sexual dimorphism, parasites were individually collected from culture by day ∼80, for blind sex assignment by bright field microscopy and confirmation by sex-specific PCR69 (Supplementary Figure S3B). Briefly, pictures from individual in vitro developed parasites were taken and sex assigned based on morphological features. Thereafter, genomic DNA from the sex-assigned individual parasites was extracted by lysing the worms at 95°C for 1h in 50mM NaOH, 0.4mM disodium ethylenediaminetetraacetic acid (EDTA), followed by the addition of neutralising buffer (80 mM Tris–HCl in water). Samples were stored at 4°C until use. PCR reactions were performed in a final volume of 10 μl consisting of 5 μl Platinum 2X hot start PCR master mix (Invitrogen, California, USA), 0.5 μl (10 μM stock) of forward and reverse primers targeting W1 repeat70 and actin71 (Supplementary Table S2), 1 μl of template DNA and 3 μl of water. The PCR program comprised an initial denature step at 94°C for 2 min followed by 30 cycles of denature at 98°C for 5 secs, and annealing/elongation at 60°C for 15 secs. The PCR products were resolved on a 1% agarose gel using 1% TAE buffer and 1kb GeneRuler DNA ladder (Thermo Scientific, UK).

Detection and quantification of proliferating cells

Parasites cultured for 15 days (D15) in media supplemented with either HS or FBS and hRBC added at D13 were harvested at days 2, 8 or 15 after cercariae transformation. Parasites cultured under the same conditions but with no hRBC were included as controls. The day before each indicated time point, a solution of 5-ethynyl-2’-deoxyuridine (EdU) was added to the culture medium at 10 μM final concentration. At each indicated time point, parasites from each group were collected and transferred to a 35 µm mesh basket (CEM Microwave Tech) in a well of a 24-well plate, washed in 1x PBS, fixed in 4% Paraformaldehyde (PFA) in 1x PBS for 30 min, washed in 1x PBS and stored at 4°C until use. The worms were incubated in PBST (1x PBS + 0.3% Triton x-100) ccontaining 2μl of Proteinase K (20 mg/ml) for 5 min at room temperature, washed three times in PBST, fixed in 4% PFA in 1x PBS for 10 min at room temperature and washed 5 times in PBST. EdU-positive cells were revealed by incubating the parasites for 30 min in the dark in a solution containing Azide fluor 488 (Sigma-Aldrich), diluted in a solution of 1 mM CuSo4, 0.025 mM Azide fluor 488, 95 mM L-Ascorbic Acid in 1x PBS. The worms were washed 5 times in PBST and incubated in mounting media containing DAPI (Fluoromount-G™ Mounting Medium, with DAPI, Invitrogen) before mounting for confocal microscopy (Leica SP8 super resolution laser confocal microscope). EdU+ cells per parasite were counted for an average of 100 parasites across three independent experiments. Statistical analysis was performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 considered significant (Supplementary Table S6).

Pairing experiments

Schistosoma mansoni adult male and female worms were collected from experimentally infected mice at 47 days post infection18. Briefly, mice were euthanised by intraperitoneal injection of 200 μl of 200 mg/ml pentobarbital supplemented with 100 U/ml heparin, and worms collected by portal perfusion (the hepatic portal vein was sectioned followed by intracardiac perfusion with phenol-red-free DMEM, containing 10 U/mL heparin) and washed in DMEM. In 24-well plates, in vitro developed female worms were cultured in the presence of in vivo developed male worms and vice versa at a ratio of 1 female every 2 males in the well containing culture medium (Table 1) supplemented with 20% HS. The parasites were cultured for several days in an incubator at 37°C, 5 % CO2 and checked daily under microscope for the presence of pairing.

Parasite staining and confocal imaging

Live parasites collected from culture at indicated time points were stained with 1:1000 dilutions of CellMaskTM Green Actin Tracking Stain (Invitrogen) and CellMaskTM Deep Red Plasma Membrane Stain (Invitrogen) to label polymerized/filamentous actin (F-actin) and cell membranes, respectively. Nuclei were stained by adding two drops per millilitre of NucBlueTM Live ReadyProbesTM Reagent (Invitrogen). After an overnight incubation at 37°C, worms were collected in 15 ml tubes, washed by gravity 3 times in 1x PBS, fixed in 4% PFA (in 1x PBS) overnight at 4°C, washed 3 times in 1x PBS and stored in 200 μl of Fluoromount-G, with DAPI (Invitrogen) before mounting on slides for confocal microscopy. All the images were acquired using a Leica SP8 super resolution laser confocal microscope.

eLife Assessment

This useful study presents an improved protocol for long-term in vitro culture of Schistosoma mansoni that enables progression toward sexually dimorphic stages, representing a meaningful advance for studying parasite development and reducing reliance on animal models. The findings show that host-specific culture conditions support essential developmental and metabolic functions required for parasite maturation, although development remains delayed compared to in vivo conditions. The evidence is solid overall, but limited pairing efficiency and the absence of egg production indicate that the system does not yet fully recapitulate complete reproductive development.

Reviewer #1 (Public review):

Pichon, Rémi et al. describe an in vitro method for transforming Schistosoma cercariae into mature adult worms. The authors show that human serum (HS) supports parasite growth and differentiation more effectively than fetal bovine serum (FBS). They also observed differences in parasite growth and activity, with worms cultured in HS efficiently digesting human red blood cells (hRBC). Cultured worms were able to pair with ex vivo adult worms and produce eggs, indicating functional maturation suitable for downstream applications such as drug screening. While the experimental approach is comprehensive and supports the advantages of HS culture conditions, the pairing efficiency was low (≈7%) and required long culture periods (70-80 days), highlighting limitations that may affect reproducibility.

A major strength of the study, in particular, is that the authors clearly differentiate the effects of FBS versus HS on developmental progression. The conversion rate observed in HS cultures is significant and consistent with previously published data.

While the study has several strengths, some aspects of the work are not fully explored. In particular, the role of hRBC supplementation requires further clarification. Although HS-cultured worms were shown to digest hRBC more readily, the implications of this observation remain unclear. Specifically, it would be useful to understand whether hRBC supplementation influences (1) long-term culture stability, (2) molecular pathways associated with development and differentiation, or (3) the pairing capacity of the worms. While addressing these questions may not be the main objective of the study, further discussion of these points would strengthen the manuscript.

The manuscript is clearly written and represents a valuable contribution to the field. Overall, the experimental approach is sound, and the results support a useful methodological framework for the in vitro culture of Schistosoma worms and the attainment of sexual maturity, particularly for adult male worms.

Reviewer #2 (Public review):

Summary:

The authors perform confirmation studies of Paul Basch's seminal schistosome work from 1981, demonstrating the development of transformed schistosomules into sexually dimorphic adult parasites, albeit without successful egg production. In addition to the findings from Basch's earlier work, the authors add some new molecular data in the form of an analysis of proliferative cells in in-vitro-derived animals.

Strengths:

The authors successfully confirm experimental results from earlier schistosome researchers, providing a potential new tool for studying schistosome biology without the need for vertebrate hosts.

Weaknesses:

The display of data from the authors is sometimes difficult to follow/understand where it comes from. For example:

(1) Line 136: The authors claim that parasites in HS and FBS conditions have substantially different mortality rates (11.3 +/- 2.7 vs 5 +/- 2.3) but a quite high p-value (0.8). Analyzing the raw data myself, I obtained a mean of 8.2 +/- 1.7% vs 4.8% +/- 4.3% with a p-value of 0.15. Either the data are not clearly presented, and I did not follow them, or the data presented in the text do not match the raw data in the supplemental files.

(2) Line 187/Figure 4: Though it is not clearly stated, it appears that the authors treat their EdU counts as an ordinal data set of 61 steps (from 0 to >60) rather than a continuous measure of EdU+ cells per animal. In this author's opinion, the graph strongly suggests a continuous data set, and the fact that this reviewer had to dig through poorly-labeled raw data to discover the nature of the data is problematic. The authors should either switch to a continuous data set or make it explicit that the data shown are ordinal. If counting EdU+ cells is too arduous, the authors could consider comparing the amount of EdU+ area to the amount of DAPI+ area in maximum intensity projections of their confocal images, as this would roughly approximate the amount of proliferative cells in the animals.

There are some minor issues as well:

(1) Line 122: It is perhaps incorrect to refer to humans as "the" definitive host of schistosomes, as S. japonicum is primarily considered a zoonotic infection with water buffalo/cows being the primary definitive host.

(2) Line 185/298: The authors refer to EdU pulse-chase experiments, but the experiments described here are EdU pulse experiments.

Reviewer #3 (Public review):

Summary:

This study is significant as it established a protocol for the long-term culture of Schistosoma mansoni newly transformed cercariae, which developed in vitro into sexually dimorphic forms. The impact of two different sera, Fetal Bovine Serum (FBS) and Human Serum (HS), added to the culture medium supplemented with human red blood cells was evaluated. The authors demonstrated that HS-cultured parasites were able to digest red blood cells, a critical step for long-term parasite development. Furthermore, while most FBS-cultured parasites did not progress beyond an early liver stage, sexual dimorphism was clearly evident in the HS-cultured worms, albeit delayed compared to in vivo development.

Strengths:

This study could contribute to further in vitro studies for a better understanding of the unique sexual biology of Schistosoma mansoni and for screening novel schistosomicidal compounds. By increasing parasite development in in vitro studies, this protocol could have a positive impact on the principles of the 3Rs (Replacement, Reduction and Refinement) for animal research.

Weaknesses:

As the authors mentioned, "pairing between male and female parasites was rare. Pairing was observed in approximately ~7% of the experiments, usually after day ~ 80 in culture. Egg production was also not achieved with this protocol.

Author response:

The following is the authors’ response to the original reviews.

eLife Assessment

This useful study presents an improved protocol for long-term in vitro culture of Schistosoma mansoni that enables progression toward sexually dimorphic stages, representing a meaningful advance for studying parasite development and reducing reliance on animal models. The findings show that host-specific culture conditions support essential developmental and metabolic functions required for parasite maturation, although development remains delayed compared to in vivo conditions. The evidence is solid overall, but limited pairing efficiency and the absence of egg production indicate that the system does not yet fully recapitulate complete reproductive development.

On behalf of the co-authors, we thank the three reviewers and the editors for their complimentary remarks as well as the major and minor comments/ concerns. Addressing these concerns have led to revisions that improved the manuscript. In particular, further analyses have generated an updated Figures 3 and 4, and Supplementary Tables S1, and S4-S6.

Public Reviews:

Reviewer #1 (Public review):

Pichon, Rémi et al. describe an in vitro method for transforming Schistosoma cercariae into mature adult worms. The authors show that human serum (HS) supports parasite growth and differentiation more effectively than fetal bovine serum (FBS). They also observed differences in parasite growth and activity, with worms cultured in HS efficiently digesting human red blood cells (hRBC). Cultured worms were able to pair with ex vivo adult worms and produce eggs, indicating functional maturation suitable for downstream applications such as drug screening. While the experimental approach is comprehensive and supports the advantage of HS culture conditions, the pairing efficiency was low (≈7%) and required long culture periods (70-80 days), highlighting limitations that may affect reproducibility.

We acknowledge the reviewer for the positive highlights. Regarding the low in vitro pairing efficiency, we have now edited the manuscript to clarify a misleading statement related to 7%. We decided to remove the value of 7% — which corresponds to the percentage of experiments in which couples were observed, as it does not accurately represent the actual number of observed worm pairs and it is probably misleading. We have updated the text as follows:

Results, lines 230 ff.:

“While the establishment of sexual dimorphism was robust and reproducible across more than 15 independent experiments, pairing between male and female parasites was rare. Pairing was observed only in experiments lasting more than 80 days in which we were only able to observe a few couples. In addition, these pairings were temporary (Figures 6A, B; Supplementary Video S4).”

We also agree with the reviewer that the extended culture periods required to obtain fully sexually dimorphic parasites remain a limitation. As elaborated in Discussion (see below), key factors, probably derived from the host, are missing in the in vitro system explaining both the slow in vitro development and low rate of spontaneous pairing between in vitro developed, sexually dimorphic male and female worms. This was discussed as follows (lines 340-343): “That said, while our system was highly efficient in producing sexually dimorphic worms, spontaneous pairing between male and female parasites was extremely rare, mainly in aged in vitro cultures (from 80 to 100 days in culture) indicating that other factors, e.g., cholesterol, may be missing [35].”

A major strength of the study, in particular, is that the authors clearly differentiate the effects of FBS versus HS on developmental progression. The conversion rate observed in HS cultures is significant and consistent with previously published data.

While the study has several strengths, some aspects of the work are not fully explored. In particular, the role of hRBC supplementation requires further clarification. Although HScultured worms were shown to digest hRBC more readily, the implications of this observation remain unclear. Specifically, it would be useful to understand whether hRBC supplementation influences (1) long-term culture stability, (2) molecular pathways associated with development and differentiation, or (3) the pairing capacity of the worms. While addressing these questions may not be the main objective of the study, further discussion of these points would strengthen the manuscript.

We agree that deciphering the role of the human Red Blood Cells (hRBCs) supplementation is critical. Regarding the influence of hRBCs on the long-term culture stability in parasite development it has been well established for more than four decades that schistosomes do need red blood cells to grow in culture [Basch, P. F. Cultivation of Schistosoma mansoni in vitro. II. production of infertile eggs by worm pairs cultured from cercariae. J Parasitol 67, 186-190 (1981); Basch, P. F. Cultivation of Schistosoma mansoni in vitro. I. Establishment of cultures from cercariae and development until pairing. J. Parasitol. 67, 179-185 (1981)]. The molecular pathways underlying development, sexual differentiation and pairing and modulated by hRBCs in culture is currently being investigated by our team. We decided not to include these data and analyses in the current manuscript, as they fall outside its scope.

The manuscript is clearly written and represents a valuable contribution to the field. Overall, the experimental approach is sound, and the results support a useful methodological framework for the in vitro culture of Schistosoma worms and the attainment of sexual maturity, particularly for adult male worms.

We thank the reviewer for highlighting the manuscript’s strengths.

Reviewer #2 (Public review):

Summary:

The authors perform confirmation studies of Paul Basch's seminal schistosome work from 1981, demonstrating the development of transformed schistosomules into sexually dimorphic adult parasites, albeit without successful egg production. In addition to the findings from Basch's earlier work, the authors add some new molecular data in the form of an analysis of proliferative cells in in-vitro-derived animals.

Strengths:

The authors successfully confirm experimental results from earlier schistosome researchers, providing a potential new tool for studying schistosome biology without the need for vertebrate hosts.

We thank the reviewer for highlighting the manuscript’s strengths.

Weaknesses:

The display of data from the authors is sometimes difficult to follow/understand where it comes from. For example:

(1) Line 136: The authors claim that parasites in HS and FBS conditions have substantially different mortality rates (11.3 +/- 2.7 vs 5 +/- 2.3) but a quite high p-value (0.8). Analyzing the raw data myself, I obtained a mean of 8.2 +/- 1.7% vs 4.8% +/- 4.3% with a p-value of 0.15. Either the data are not clearly presented, and I did not follow them, or the data presented in the text do not match the raw data in the supplemental files.

We thank the reviewer for pointing this out; we have now edited Supplementary Tables S1 and S6 by turning them into a long format for the sake of clarity. Accordingly, Results, Methods sections, and indicated supplementary tables were edited as follows:

Results, lines 142 ff.:

“No morphological differences were observed between parasites cultured either in FBS or HS within the first week in culture; in both conditions most parasites were classified as early schistosomula [category 1: 76% ± 30 (average ± SD) in FBS and 73% ± 29 (average ± SD) in HS] with few lung (category 2) and early liver schistosomula (category 3) (Figure 1B, week 1; Supplementary Figure S1). The mean mortality (category 0) at week 1 was slightly higher, but not statistically significant (P= 0.42), in worms cultured in HS [9.75% ± 2.76 (average ± SD)] compared to the mortality registered in FBS-cultured parasites [5.52% ± 5.18 (average ± SD), Supplementary Table S6], consistent with previous findings [39].”

Methods, lines 463-465:

“To evaluate differences in mortality between HS- and FBS-cultured parasites, data from 5 experiments were combined and analysed using a Shapiro-Wilk normality test to test normality of the data and a non-parametric Wilcoxon rank sum exact test (Supplementary Tables S1 and S6).”

Supplementary Tables:

Supplementary Table S1. “Raw counts of parasites within each developmental stage category. Each row corresponds to a picture of parasites in culture medium containing FBS or HS. Each column corresponds to the raw parasite counts at indicated stage development (categories 0 to 5), time in culture (Time in days - D), and experimental condition.”

Supplementary Table S6. “Summary of all statistical tests employed in this study. 1. Statistical tests of parasite mortality and the raw data table used for this test. 2. Statistical tests for worm size comparisons (correspond to Figure 2). 3. Statistical tests for worm black gut comparisons (correspond to Figure 3). BG: Black gut. 4. Statistical tests for EdU positive cells comparisons (correspond to Figure 4). Replicate code: E, M and L correspond to day 2, 8 and 15 respectively; R and W correspond to the presence (R) or absence (W) of RBCs added 13 days after transformation.”

For clarity, below we provide the R script used to perform the statistical tests on the data shown in Supplementary Table S6 (column ‘Raw count of parasite developmental category per image and experiment’)

(2) Line 187/Figure 4: Though it is not clearly stated, it appears that the authors treat their EdU counts as an ordinal data set of 61 steps (from 0 to >60) rather than a continuous measure of EdU+ cells per animal. In this author's opinion, the graph strongly suggests a continuous data set, and the fact that this reviewer had to dig through poorly-labeled raw data to discover the nature of the data is problematic. The authors should either switch to a continuous data set or make it explicit that the data shown are ordinal. If counting EdU+ cells is too arduous, the authors could consider comparing the amount of EdU+ area to the amount of DAPI+ area in maximum intensity projections of their confocal images, as this would roughly approximate the amount of proliferative cells in the animals.

As the reviewer correctly pointed out, the data were treated as ordinal because counting worms with more than 60 Edu+ cells became extremely difficult and highly inaccurate. Therefore, we decided to group in a single category, “60 EdU+ cells”, all worms showing more than 60 EdU+ cells. We have now updated Figure 4 where medians are shown instead of media values, Supplementary Table S5 to provide more comprehensive access to the raw counts, and Supplementary Table S6 to indicate the data for EdU+ cells per worm were considered ordinal. Accordingly, we have revised the corresponding sections as follows:

Results, lines 211 ff:

“HS-cultured schistosomula showed higher numbers of proliferating stem cells, with a median of >48 and >60 EdU+ cells per worm at days 8 and 15, respectively (Figure 4). On the other hand, most FBS-cultured parasites displayed no more than an average of 20 EdU+ cells per worm (Figure 4).”

Methods, lines 520 ff:

“EdU+ cells per parasite were counted for an average of 100 parasites across three independent experiments (Supplementary Table S5). Worms were grouped based on the number of cells per individual, but all those showing ⪰ 60 EdU+ cells were counted in the same group named ‘60 EdU+ cells'. Therefore, the data were considered ordinal data. Statistical analysis was performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 considered significant (Supplementary Table S6).”

Figure 4 legend, lines 830 ff:

“A. Violin plots showing the number of Edu+ cells per worm at indicated time points (2, 8, and 15 days post cercarial transformation) in parasites cultured either in Foetal Bovine Serum (FBS, blue) or Human Serum (HS, light brown). Human Red Blood Cells (hRBCs) were added in the culture at day 13 post cercarial transformation. The small black dots indicate individual worms, and the big black point indicates the median of EdU+ cells per worm. All worms showing ⪰ 60 EdU+ cells were counted and clustered together in the group named ‘60 EdU+ cells’. Hence, the data were treated as ordinal and statistical analysis performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 (*) considered significant (Supplementary Tables S5 and S6).”

We thank the reviewer for the very interesting suggestion to quantify cell proliferation by calculating the ratio between EdU+ area to DAPI+ area in maximum intensity projections images. Measuring the fluorescence area for each worm in maximum projection is an excellent idea; however, due to the number of EdU+ cells present in some samples, we think this technique would not provide additional information or produce more detailed data compared with our analysis when the number of Edu+ cells exceeds 60 per worm. We will certainly consider this approximation for future studies.

There are some minor issues as well:

(1) Line 122: It is perhaps incorrect to refer to humans as "the" definitive host of schistosomes, as S. japonicum is primarily considered a zoonotic infection with water buffalo/cows being the primary definitive host.

We thank the reviewer for pointing this out; we have now replaced ‘schistosomes’ with ‘Schistosoma mansoni’ (current line 131)

(2) Line 185/298: The authors refer to EdU pulse-chase experiments, but the experiments described here are EdU pulse experiments.

This is a very good point, we thank the reviewer for bringing this up and have accordingly edited by replacing ‘EdU pulse-chase’ with ‘EdU pulse’ experiments in lines 37, 204, and 321.

Reviewer #3 (Public review):

Summary:

This study is significant as it established a protocol for the long-term culture of Schistosoma mansoni newly transformed cercariae, which developed in vitro into sexually dimorphic forms. The impact of two different sera, Fetal Bovine Serum (FBS) and Human Serum (HS), added to the culture medium supplemented with human red blood cells was evaluated. The authors demonstrated that HS-cultured parasites were able to digest red blood cells, a critical step for long-term parasite development. Furthermore, while most FBS-cultured parasites did not progress beyond an early liver stage, sexual dimorphism was clearly evident in the HS-cultured worms, albeit delayed compared to in vivo development.

Strengths:

This study could contribute to further in vitro studies for a better understanding of the unique sexual biology of Schistosoma mansoni and for screening novel schistosomicidal compounds. By increasing parasite development in in vitro studies, this protocol could have a positive impact on the principles of the 3Rs (Replacement, Reduction and Refinement) for animal research.

We thank the reviewer for highlighting the manuscript’s strengths.

Weaknesses:

As the authors mentioned, "pairing between male and female parasites was rare. Pairing was observed in approximately ~7% of the experiments, usually after day ~ 80 in culture. Egg production was also not achieved with this protocol.

Following the reviewer’s point and to clarify a misleading point, we have now decided to remove the value of 7% - which corresponds to the percentage of experiments in which couples were observed. However, this value does not accurately reflect the actual number of observed worm pairs, and it is probably misleading. We have updated the text as follows:

Results, lines 230 ff:

“While the establishment of sexual dimorphism was robust and reproducible across more than 15 independent experiments, pairing between male and female parasites was rare. Pairing was observed only in experiments lasting more than 80 days in which we were only able to observe a few couples. In addition, these pairings were temporary (Figures 6A, B; Supplementary Video S4).”

Recommendations for the authors:

Reviewer #1 (Recommendations for the authors):

The manuscript is well-written overall. However, there are some minor revisions that would further improve the clarity and presentation of the data.

(1) At the beginning of the manuscript, it would be helpful to clearly state three to four specific aims or objectives. This would help readers better understand the expected outcomes and the broader methodological contribution of the study.

We agree with the reviewer and accordingly have stated the overall goals of the study, as follows:

Introduction, lines 106 ff:

“We aimed at optimising a platform to study intra-mammalian schistosomes that supports in vitro sexual dimorphism establishment, consequently leading to an overall positive impact in the 3Rs (Reduction, Replacement, Refinement) for animal research (https://nc3rs.org.uk/) [42]”.

(2) In the abstract, you highlighted the relevance of the work according to the 3R principles of reduction in animal experimentation. However, this point is not clearly introduced in the Introduction section. Including a short discussion of this aspect would improve continuity and context.

Following this and previous item raised by the reviewer, we have now clarified the potential impact in the 3Rs by our research outcomes and included that link to the NC3Rs website and a representative reference [Louis-Maerten E, Rodriguez Perez C, Cajiga RM, Persson K and Elger BS (2024). Conceptual foundations for a clarified meaning of the 3Rs principles in animal experimentation. Animal Welfare, 33, e37, 1–11)].

(3) In line 43, please italicize Schistosoma spp.

Edited accordingly.

(4) When discussing the importance of "interfering with sexual development," in line 52, please specify the life cycle stages being referred to.

Revised accordingly as follows:

Introduction, lines 54-56:

“This suggests that interfering with the sexual development of schistosome intra-mammalian stages could potentially restrict human pathology.”

(5) Between lines 56-58, please rephrase this sentence for clarity.

We thank the reviewer for this editorial suggestion. The text has been revised as follows:

Introduction, lines 58 ff :

“Therefore, novel control strategies are urgently needed, and new targets for drug/ vaccine development became a priority. A better understanding of the mechanisms underlying schistosome development, including sexual dimorphism establishment, will pave the wave to achieve this goal.”

(6) In lines 66-68 & line 88, please clarify whether the transcriptomic studies cited were performed in vivo, in vitro, or ex vivo, and indicate the developmental stages analyzed.

We have now included the information suggested by the reviewer as follows:

Introduction, lines 69-70:

“Transcriptomic studies, at both bulk [7-11] and single cell [12-1]4 levels for intra mammalian stages in vivo and ex vivo,...”

(7) Please indicate, in line 110, the day of culture for reference. Without this information, the conversion rates per life cycle stage are difficult to interpret and reproduce. Overall, please try to give an overview in the text of these rates of conversion for context, wherever possible.

Following the reviewer’s question, we have clearly indicated the in vitro and in vivo timings for ‘conversion’ (understood as sexual dimorphism establishment.) We have written:

Introduction, lines 117-120:

“Finally, while most of the FBS-cultured parasites did not progress beyond lung and early liver stage, HS-cultured parasites reached sexually dimorphic stages by week 6, albeit at a slightly delayed rate compared to in vivo development. In the mouse model, parasites become dimorphic by day 21 post-infection (~3 weeks) [12].”

(8) The section beginning with "Furthermore, phenotypic...cell proliferation" (line 110) may be easier to follow if moved earlier in the Introduction.

Following the reviewer’s suggestion, we have moved and slightly rewritten the sentence to current line 112, as follows: “First, phenotypic differences between FBS- and HS- cultured parasites became evident as early as 48 hours in culture, with HS-cultured parasites exhibiting higher rates of cell proliferation resulting in larger worms in the HS condition.”

(9) In line 126, please remove the DOI and add the citation.

Edited accordingly.

(10) When referring to 10-week-old parasites, in line 130, please indicate the developmental stage at which they stalled and relate this to the phenotypic scoring shown in Figure 1.

Based on this suggestion, we have now revised the third paragraph of Results section (‘Sexually dimorphic schistosomes developed entirely in vitro from cercariae’), as follows:

Results, lines 137 ff.:

“The development of schistosomula derived from mechanically transformed cercariae was assessed in at least 15 independent experiments, five of which were maintained over a period of at least 10 weeks to assess parasite survival and ability to mate and produce fertile eggs (Figure 1A; Supplementary Table S1).”

Lines 151 ff.:

“Differences in parasite development between the two conditions became apparent by week 2 (Figure 1B). At this time point, 14.8% ± 24.9 (average ± SD, excluding dead worms) or 36% ± 33.6 (average ± SD, excluding dead worms) of the parasites cultured in FBS or HS, respectively, have reached category 3, i.e., early liver schistosomulum. Parasites in FBS rarely progressed beyond this stage during the 10-week experiment, with very few parasites (<0.1% ± 0.2, average ± SD) reaching category 4, i.e., late liver schistosomulum. In contrast, worms cultured in HS developed over time across all categories, achieving marked sexual dimorphism by week 6 (13.4% ± 18.6, average ± SD) (Figure 1B; Supplementary Figure S3A), as confirmed by PCR (Supplementary Figure S3B; Supplementary Table S2). No differences in the timing for sexual dimorphism establishment were observed between male and female parasites. The mortality rate of FBS-cultured parasites reached an average of 76.24% ± 23.46 (average ± SD) by week 10, after which the experiments under this condition were stopped as most parasites were dead (Supplementary Figure S2). From that time point onwards only parasites in HS were kept in culture. As previously described for the in vivo development of schistosomes [12], in vitro cultured parasites showed developmental asynchrony in agreement with Basch’s observations [33]; however, by week 10 most of the worms in HS (73.7% ± 25.4, average ± SD) acquired an evident sexual dimorphism (Figure 1B).”

(11) In line 142, please provide a standard deviation value for the reported average of 14.8%, if available. As well as the absolute numbers of these parasites or indicate them in the supplementary. Otherwise, it is difficult to understand the true conversion rate.

We followed the reviewer’s suggestions and have now rewritten the text (see above, item 10). In addition, Supplementary Table S1 was edited in long format (see answer for item 1, reviewer #2)

(12) Please explain, IN line 144, why all cultures were maintained for 10 weeks and provide the rationale for this experimental design.

We thank the reviewer for this opportunity to clarify this point and hence improve the manuscript. The experimental condition stopped at week 10 included only FBS-cultured worms, not HS-cultured parasites. This is relevant as most of the parasites in FBS were dead by this time, unlike the HS-developed schistosomes. Indeed, some experimental groups consisting of parasites cultured in HS were maintained for up to 22 weeks. We have now updated the text to clarify this point, as follows:

Results, lines 160 ff.:

“The mortality rate of FBS-cultured parasites reached an average of 76.24% ± 23.46 (average ± SD) by week 10, after which the experiments under this condition were stopped as most parasites were dead (Supplementary Figure S2). From that time point onwards only parasites in HS were kept in culture.”

(13) In lines 146-151, please streamline the timelines of culture conditions and observed outcomes in FBS versus HS media. As the current wording makes interpretation difficult.

Following the reviewer’s suggestion we have streamlined the culture timelines and observed outcomes, as follows:

Results, lines 137 ff.:

“The development of schistosomula derived from mechanically transformed cercariae was assessed in at least 15 independent experiments, five of which were maintained over a period of at least 10 weeks to assess parasite survival and ability to mate and produce fertile eggs (Figure 1A; Supplementary Table S1).”

Results, lines 151 ff.:

“Differences in parasite development between the two conditions became apparent by week 2 (Figure 1B). At this time point, 14.8% ± 24.9 (average ± SD, excluding dead worms) or 36% ± 33.6 (average ± SD, excluding dead worms) of the parasites cultured in FBS or HS, respectively, have reached category 3, i.e., early liver schistosomulum. Parasites in FBS rarely progressed beyond this stage during the 10-week experiment, with very few parasites (<0.1% ± 0.2, average ± SD) reaching category 4, i.e., late liver schistosomulum. In contrast, worms cultured in HS developed over time across all categories, achieving marked sexual dimorphism by week 6 (13.4% ± 18.6, average ± SD) (Figure 1B; Supplementary Figure S3A), as confirmed by PCR (Supplementary Figure S3B; Supplementary Table S2). No differences in the timing for sexual dimorphism establishment were observed between male and female parasites. The mortality rate of FBS-cultured parasites reached an average of 76.24% ± 23.46 (average ± SD) by week 10, after which the experiments under this condition were stopped as most parasites were dead (Supplementary Figure S2). From that time point onwards only parasites in HS were kept in culture. As previously described for the in vivo development of schistosomes [12], in vitro cultured parasites showed developmental asynchrony in agreement with Basch’s observations [33]; however, by week 10 most of the worms in HS (73.7% ± 25.4, average ± SD) acquired an evident sexual dimorphism (Figure 1B).”

(14) In lines 153-159, please clarify comparisons between worms cultured in FBS and HS at equivalent time points (e.g., 2 weeks FBS vs 2 weeks HS), rather than comparing only 10 week cultures.

Following the reviewer’s comment, we have now rewritten the whole third paragraph in Results, under the heading “Sexually dimorphic schistosomes developed entirely in vitro from cercariae” - changes detailed in answers to items 10 and 13 (above).

(15) It would also be helpful to include information on male versus female development in the context of sexual dimorphism.

This is a relevant point that we have not clarified in the original submission - we have now indicated in the text that no differences were detected in the timing for male and female dimorphism establishment. New text included as follows:

Results, lines 159-160:

“No differences in the timing for sexual dimorphism establishment were observed between male and female parasites.”

(16) In line 163, please resolve the editing marks and punctuation.

Resolved accordingly.

(17) In lines 169 and 172, when referring to stages such as "early liver stage," please indicate the corresponding time in culture (e.g., 3 weeks, 7 weeks + 3 days), or define these stage classifications earlier in the manuscript.

Following the reviewer’s suggestion we have now included the developmental category after stating ‘early liver stage’, as follows:

Results, line 187:

“Even though few parasites in FBS reached the early liver stage (category 3)…”

(18) Please indicate, in line 173, the developmental stage of worms used when assessing hRBC digestion in HS and FBS cultures. Additionally, here, it would be useful to discuss how hRBC supplementation may influence worm development beyond culture conditions, including possible molecular mechanisms. As a revision, that way maybe you can include data, if already performed or conduct it, to show the effect of adding or not adding hRBC even in HS cultured worms.

We thank the reviewer for highlighting this important item that warrants further clarification. As stated in Results washed human red blood cells (hRBCs) were added to the culture at day 13. Pilot experiments in which hRBCs were added at different time points had been previously performed; no hemoglobin digestion was apparent when hRBCs were added at days 4, 5 and 6 consistent with previous findings (Correnti JM, Jung E, Freitas TC, Pearce EJ. Transfection of Schistosoma mansoni by electroporation and the description of a new promoter sequence for transgene expression. Int J Parasitol. 2007 Aug;37(10):1107-15. doi: 10.1016/j.ijpara.2007.02.011. Epub 2007 Mar 18. PMID: 17482194.).

Following this observation, we have added a line to clarify this point, as follows (lines 181187): “Based on both previous reports [45], and pilot experiments in which adding human Red Blood Cells (hRBCs) to the culture before day ~10 did not show obvious haemoglobin digestion, we decided to supplement the culture media with hRBCs at day 13. The addition of hRBCs allowed the parasites to feed and thus continue their development [19]. At this point, they began to swallow and degrade erythrocytes, producing hemozoin, a black pigment derived from host haemoglobin degradation and visible in the worms' intestines.”

Regarding the specific effect of adding hRBCs in the culture, this is a very good point. First, it has been well established for more than four decades that schistosomes need red blood cells in culture to grow, as example see (Basch, P. F. Cultivation of Schistosoma mansoni in vitro. II. production of infertile eggs by worm pairs cultured from cercariae. J Parasitol 67, 186-190 (1981); Basch, P. F. Cultivation of Schistosoma mansoni in vitro. I. Establishment of cultures from cercariae and development until pairing. J. Parasitol. 67, 179-185 (1981). Second, we are currently analysing transcriptomic data from parasites cultured in different conditions, including in the presence or absence of hRBCs. We decided not to include these data and analyses in the current manuscript, as they fall outside its scope.

(19) In line 183, please clarify whether the referenced single-cell transcriptomic data were obtained from adult worms.

We have now clarified this point in the manuscript as follows:

Results, lines 199 ff:

“In schistosomes, a complex stem cell system consisting of both somatic and germline stem cells has been described by leveraging recent single cell transcriptomic data across different developmental stages, including schistosomula and adult worms [47].”

(20) In lines 210 and 213, please indicate the absolute number of worms used for these observations, rather than only percentages. If possible, also report any sex bias in pairing.

Following this and a similar item raised by reviewer #3 (public review), we decided to remove the mention of 7% given it is misleading. This percentage corresponds to the percentage of experiments in which couples were observed. However, this value does not accurately reflect the actual number of observed worm pairs, and it is probably misleading. We have updated the text as follows:

Results, lines 230 ff.:

“While the establishment of sexual dimorphism was robust and reproducible across more than 15 independent experiments, pairing between male and female parasites was rare. Pairing was observed only in experiments lasting more than 80 days in which we were only able to observe a few couples. In addition, these pairings were temporary (Figures 6A, B; Supplementary Video S4).”

(21) In the final results section, please clarify whether pairing enhances sexual maturation of already mature worms or whether maturation occurs primarily after pairing.

This is a very relevant point, and we thank the reviewer for giving us the opportunity to clarify it in the manuscript. As described in the manuscript the parasite sexual dimorphism was established in vitro and developed male and female parasites were capable of pairing. Moreover, enlarged oocytes in the ovary’s posterior section of in vitro developed female parasites became apparent after pairing. This observation (Figure 6E, F and Supplementary Video S6) suggests that these female parasites, fully developed in HS-supplemented culture media, were not only capable of pairing, but of starting to fully maturate. We have clarified this aspect in the manuscript as follows:

Results, lines 243 ff.:

“Moreover, in vitro developed females coupled with ex vivo collected mature males displayed signs of primordial ovary maturation with larger oocytes towards the posterior region of the ovary (Figure 6E, F; Supplementary Video S6). On the other hand, females developed in vitro but not paired with ex vivo collected males remained immature.”

(22) Further in the Materials and methods sections, please clarify, isn't 8000 schistosomula/well of a 6-well plate really a confluent culture condition, and does it contribute to NTS mortality in that way, as shown in previous in vitro transformation publications? Please clarify, at least with relative values, percentages of parasite transformation in such a concentrated system.

No formal titration experiments were carried out but based on empirical observations during pilot experiments we decided to add no more than 8,000 schistosomula per well. This is something to further investigate in the future. We have now added the following sentence in Methods:

Methods, lines 423-426:

“The number of parasites cultured per well (~8,000 schistosomula) was determined empirically, as no formal titration experiments were performed. At higher densities (>10,000 per well), more frequent media changes were required, and parasite development appeared to be impaired.”

(23) Also, what was the rationale of adding hRBCs as early as 13 days post-transformation, when the parasites are in the lung and early liver stage, just forming the guts? Therefore, is it possible that this would have contributed to the observation of lesser parasites disgesting hRBCs? Also, were the hRBC supplemented each time with the media change? This was not clear.

We thank the reviewer for these questions. The rationale of adding hRBCs at day 13 has been elaborated above (question 18). In addition, in the mouse model, parasites have already migrated through and left the lungs by day 13 post-infection, as described by Nation et al [Nation CS, Da’dara AA, Marchant JK, Skelly PJ (2020) Schistosome migration in the definitive host. PLoS Negl Trop Dis 14(4): e0007951] as follows: “In the mouse, S. mansoni schistosomula begin to arrive in the lungs between 2 and 3 days post-infection, peaking at around day 7 and lasting until around day 11”. Hence, we do not think that adding hRBCs at day 13 contributed to the observation of fewer parasites digesting hemoglobin, because this was only seen in parasites cultured in FBS, not in HS.

The hRBCs were replaced every two weeks, or sooner if their numbers decreased due to consumption. We have now clarified this point in Methods as follows (lines 427-430): “LTC medium was replaced twice a week and washed human red blood cells (hRBCs) added to a final concentration of 0.02% v/v at 13 days after transformation. Washed hRBCs were replaced every two weeks, or sooner if their numbers decreased due to consumption.”

(24) In the Discussion, please address the limitations related to the relatively late onset and low frequency of pairing in vitro.

Following the reviewer’s suggestion and comments from reviewer #1, we have now included a section in Discussion highlighting the limitations of the study and avenues to overcome these in the future.

Discussion, line 360 ff.:

“Considering these elements in future experiments will help overcome the limitations encountered in this study, including the low rate of spontaneous pairing between in vitro– developed male and female worms and the requirement for extended culture periods (>70 days). In addition, further research is needed to assess the role of host- and parasite-derived cues in schistosome development.”

(25) Figure 1: Please consider adding arrows or markers indicating which parasites correspond to the representative developmental stages used for classification.

We acknowledge the reviewer for the suggestion; however, we respectfully consider this may not be necessary as (1) the images shown in Figure are representative pictures of each time point included for illustrative purposes; (2) Supplementary Figure S1 clearly depicts representative images of worms in each developmental category associated with specific morphological descriptions. For greater clarity we have now added the following text at the end of Figure 1 legend:

Figure 1 legend, line 810-811:

“A detailed description of the developmental categories and representative images are provided in Supplementary Figure S1.”

(26) Figure 2: This plot is somewhat misleading in showing that the HS cultured worms grew significantly more than the FBS worms, where the latter did not grow at all, as also shown by the blue bars all over the plot.

We appreciate the reviewer’s observation; critically, the data shown in Figure 2 represent measurements of the worm's area, which means that some worms may have become longer but thinner maintaining the same area. Most of the FBS-cultured worms did not develop beyond lung or early liver stages, in which the parasites were long/ thin or shorter/wide, respectively. Therefore, the overall area of these FBS-cultured worms almost did not change (please see the raw data and statistical analyses in Supplementary Tables S3 and S6. We believe that, as presented, Figure 2 is sufficiently clear and self-explanatory. However, we would be happy to consider any suggestions to further clarify this point in the manuscript.

(27) Figure 3: For panel A, what is the worm percentage corresponding to? The context is missing. Please clarify in the text.

Following the reviewer’s question and for clarity, we have now (1) modified the axis-legend in Figure 3 as “Percentage of worms displaying or not Black Guts - BG (%)”, and (2) slightly edited the legend as follows:

Figure 3 legend, lines 820-823:

“Bar Plot representing the percentage of Human Serum (HS)- or Foetal Bovine Serum (FBS)-cultured schistosomula with (blue bar) or without (light brown bar) black guts (BG) due to the presence of intestinal hemozoin.”

Reviewer #2 (Recommendations for the authors):

The authors need to clarify their presentation of data. The raw data needs to be more clearly labeled/explained, and the representation of the data in Figure 4A needs to be explicitly described or changed.

We acknowledge the reviewer for highlighting this issue related with the data presentation and have decided to follow their advice by editing Figures 3 and 4, and improving the data presentation in Supplementary Tables S1, and S4-S6. In particular:

Figure 3. We have now modified the axis-legend as “Percentage of worms displaying or not Black Gut - BG (%)”, and slightly edited the legend as follows:

Figure 3 legend, lines 820-823:

“Bar Plot representing the percentage of Human Serum (HS)- or Foetal Bovine Serum (FBS)-cultured schistosomula with (blue bar) or without (light brown bar) black guts (BG) due to the presence of intestinal hemozoin.”

Figure 4. We have edited this figure to show medians instead of media values, and updated the legend as follows: lines 830 ff.:

“A. Violin plots showing the number of Edu+ cells per worm at indicated time points (2, 8, and 15 days post cercarial transformation) in parasites cultured either in Foetal Bovine Serum (FBS, blue) or Human Serum (HS, light brown). Human Red Blood Cells (hRBCs) were added in the culture at day 13 post cercarial transformation. The small black dots indicate individual worms, and the big black point indicates the median of EdU+ cells per worm. All worms showing ⪰ 60 EdU+ cells were counted and clustered together in the group named ‘60 EdU+ cells’. Hence, the data were treated as ordinal and statistical analysis performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 (*) considered significant (Supplementary Tables S5 and S6).”

Supplementary Table S1. We have clarified the data presentation by turning it into a long format and updated the legend accordingly as follows (lines 864-867): “Raw counts of parasites within each developmental stage category. Each row corresponds to a picture of parasites in culture medium containing FBS or HS. Each column corresponds to the raw parasite counts at indicated stage development (categories 0 to 5), time in culture (Time in days - D), and experimental condition.”

Supplementary Table S4. We have clarified the table by turning it into a long format, simplified the data presentation, and updated the legend accordingly as follows (lines 873874): “Percentage of parasites displaying either black positive (hemozoin) or black negative (no hemozoin) intestine.”

Supplementary Table S5. We have simplified the table by turning it into a long format, and explained the naming for elements in columns C (‘Group’) and D (‘Replicate’). We have updated the legend accordingly as follows (line 876 ff.): “Raw counting of EdU positive cells per parasite for indicated experimental group, replicate and experiment in long format. The worms were classified by group (column C) and replicate (column D), using the following code: E (‘early’), M (‘medium’) and L (‘late’), corresponding to days 2, 8 and 15, respectively. R and W correspond to conditions with (R) or without (W) human red blood cells, and HS and FBS to culture medium employed.”

Supplementary Table S6. We have incorporated a new section with the statistical analyses for parasite mortality estimation and updated the legend accordingly as follows (lines 882887): “Summary of all statistical tests employed in this study. 1. Statistical tests of parasite mortality and the raw data table used for this test. 2. Statistical tests for worm size comparisons (correspond to Figure 2). 3. Statistical tests for worm black gut comparisons (correspond to Figure 3). BG: Black gut. 4. Statistical tests for EdU positive cells comparisons (correspond to Figure 4). Replicate code: E, M and L correspond to day 2, 8 and 15 respectively; R and W correspond to the presence (R) or absence (W) of RBCs added 13 days after transformation.”

Reviewer #3 (Recommendations for the authors):

The study was well conducted, and the data presented clearly support the conclusions. The protocol is well described, making it reproducible. The pairing experiments could be improved.

Specific Questions.

(1) "Male and female adult worms that developed in vivo and recovered from mice by portal perfusion on day 42 post-infection were sorted by sex and placed in culture with worms of the opposite sex developed in vitro (>70 days). Within 24 hours of initiating the co-culturing of in vitro developed worms with ex vivo collected worms, couples were observed".

In the interest of clarity, and considering that stating ‘worms developed in vivo were collected from infected mice’ is redundant, we have now shortened and edited these lines as follows (lines 238- 242): “Male and female adult worms were recovered from mice by portal perfusion on day 42 post-infection, sorted by sex and placed in culture with worms of the opposite sex developed in vitro. Within 24 hours of initiating the co-culturing of in vitrodeveloped worms with ex vivo collected worms, couples were observed (Figures 6C, D; Supplementary Video S5).”

(2) Have the authors conducted experiments with in vitro female and male parasites under the same experimental conditions as the in vitro/ex vivo pairing experiments? Is it possible that the tissue culture medium used for the development of sexually dimorphic forms is inhibiting pairing?

The reviewer raises an interesting point that warrants clarification. First, the experimental conditions tested for in vitro developed parasites were the same as for the pairing experiments, as the ex vivo collected worms were washed and placed in HS-supplemented media. Second, as the culture conditions were the same (same culture protocol and medium) between in vitro pairing and in vitro / ex vivo pairing experiments, we do not think that the tissue culture medium used for developing sexually dimorphic parasites inhibited the pairing. As elaborated in Discussion (see below), key factors, probably derived from the host, are missing in the in vitro system explaining the low rate of spontaneous pairing between in vitro developed, sexually dimorphic male and female worms. This was discussed as follows (lines 340-343): “That said, while our system was highly efficient in producing sexually dimorphic worms, spontaneous pairing between male and female parasites was extremely rare, mainly in aged in vitro cultures (from 80 to 100 days in culture) indicating that other factors, e.g., cholesterol, may be missing [35].”

Version 2

Abstract

Schistosomes are parasitic flatworms that cause Schistosomiasis, a major Neglected Tropical Disease (NTD) that affects more than 250 million people worldwide. With two morphologically distinct sexes, a heterogametic female (ZW) and a homogametic male (ZZ), schistosomes are an exception among flatworms, which are largely hermaphroditic. Sexual dimorphism in schistosomes only becomes apparent by adulthood within the mammalian host. The cellular and molecular mechanisms underlying the sexual differentiation of these parasites are poorly understood, partly due to intrinsic challenges in assessing parasite development in vivo. Therefore, robust and reproducible approaches for maintaining and developing parasites in vitro are needed to overcome these difficulties. However, to date, only a few studies have focused on protocols that allow cultured parasites to reach sexual dimorphic stages, and none have been reproduced, limiting the ability to understand the unique sexual biology of this major human parasite. Here, we refine a protocol for long-term culture of newly transformed cercariae that developed in vitro into sexually dimorphic forms. We assessed the effect of two different sera, Foetal Bovine Serum (FBS) and Human Serum (HS), added to the culture medium supplemented with human red blood cells. We found that in contrast to FBS-cultured worms, HS-cultured parasites digested red blood cells, a crucial step for long term parasite development. Additionally, while most FBS-cultured parasites did not progress beyond an early liver stage, sexual dimorphism was clearly established in the HS-cultured worms, albeit delayed compared to in vivo development. Moreover, EdU pulse experiments revealed a continuous proliferation of cells over time in HS-cultured parasites, while a significantly lower number of proliferating cells were detected in FBS-cultured worms. This protocol paves the way to study parasite development in vitro, positively impacting the principles of the 3Rs (Replacement, Reduction and Refinement) for animal research, and allowing for in-depth studies of sexual dimorphism establishment as well as in vitro screening for novel control strategies across the life cycle of these major human parasites.

Introduction

Schistosomiasis, a major Neglected Tropical Disease (NTD) affecting >250 million people worldwide, is caused by the infection with blood flukes (class Trematoda) in the genus Schistosoma1. The pathology associated with schistosomiasis is largely driven by egg trapping in different tissues depending on the species. Infection with Schistosoma mansoni leads to eggs lodged mainly in the liver, inducing inflation, fibrosis and granuloma formation2. This suggests that interfering with the sexual development of schistosome intra-mammalian stages could potentially restrict human pathology. Currently, the complete reliance on a single drug (Praziquantel) to treat the infection and its use in drug mass administration programs in endemic areas threatens the development of drug resistance3,4. Therefore, novel control strategies are urgently needed, and identifying original targets for drug/ vaccine development became a priority. A better understanding of the mechanisms underlying schistosome development, including sexual dimorphism establishment, will pave the wave to achieve this goal.

While most parasitic flatworms are hermaphrodites, schistosomes are dioecious with genetically determined female (2n=16, ZW) and male (2n=16, ZZ) individuals. Whilst male and female parasites are morphologically indistinguishable across most developmental stages, sexual dimorphism becomes apparent by adulthood within the mammalian host5. Male and female worms undergo separate but concurrent sexual differentiation of their gonads and somatic tissues that eventually allows intersexual pairing, a critical step for female maturation, egg production and life cycle propagation6,7. Transcriptomic studies, at both bulk8–12 and single cell13–15 levels for intra mammalian stages in vivo and ex vivo, have revealed key pathways and molecular mediators underlying the critical process of female sexual maturation induced by pairing. However, the cellular and molecular mechanisms driving sexual dimorphism prior to pairing remain poorly characterised, partly due to intrinsic challenges in assessing the development of the parasite in vivo12.

Approaches to better understand the biology of parasites at large have been developed and optimised using in vitro or ex vivo refined culture systems16,17. Combinations of culture media components, sera from different sources and additives have been tested18,19. In addition, organ-on-a-chip systems and organoid-based 2D and 3D culture platforms20,21 have first been implemented for protozoa parasites such as Plasmodium, Toxoplasma, and Cryptosporidium species. These methods have facilitated the molecular dissection of processes underlying single-cell parasite development and interaction with the host22–24. Recently, these technologies have started to be transferred to metazoan parasites to study host-parasite interaction and immune response in the context of helminth infections25–29. In vitro cultivation of both larval tapeworms (cestodes) and stem cells isolated from cestodes, coupled with ‘omic’ and imaging analyses have unveiled the role of stem cells in parasite development30,31. However, despite these significant developments in the field, reproducing the natural development of trematodes under controlled in vitro conditions remains extremely challenging17.

Significant progress was achieved in culturing schistosomes when, more than forty years ago, Paul F. Basch developed and optimised comprehensive culture protocols for intra-mammalian developmental stages of schistosomes32–34. Basch reported for the first time the development of male and female parasites from cercariae, with worm pairing and production of infertile eggs entirely in vitro32. Remarkably, to the best of our knowledge, these results have not been replicated and no further attempt to obtain egg-laying worm pairs developed in vitro from cercariae has been reported since Basch’s seminal studies17. More recently, in vitro culture protocols for schistosomes have been refined either to maintain ex vivo parasites collected from infected mice to study parasite reproduction biology35,36, or to culture juvenile parasites from cercariae and develop drug screening approaches37–39. However, no long-term culture conditions to specifically assess schistosome sexual differentiation have been reported since 198133. Despite one research group referring to the approach40, there has been no detailed description and no widespread adoption. In fact, published culturing studies have continued to utilise FBS35,41.

Novel functional approaches applied to culture systems that allow reliable and reproducible in vitro establishment of sexual dimorphism will shine new light into molecular mechanisms underlying schistosome intra-mammalian development. We aimed at optimising a platform to study intra-mammalian schistosomes that supports in vitro sexual dimorphism establishment and consequently leading to an overall positive impact in the 3Rs (Reduction, Replacement, Refinement) on animal research (https://nc3rs.org.uk/)42. Here, we refined a protocol for long-term culture of newly transformed cercariae by assessing the effects of two different sera, Foetal Bovine Serum (FBS) and Human Serum (HS), added to the medium supplemented with human Red Blood Cells (hRBCs). Striking differences were evident between parasites maintained in either FBS or HS. First, phenotypic differences between FBS- and HS- cultured parasites became evident as early as 48 hours in culture, with HS-cultured parasites exhibiting higher rates of cell proliferation resulting in larger worms in the HS condition. Second, hRBC were digested and hemozoin became apparent in the intestines of HS-cultured parasites within a day in contrast to FBS-cultured parasites that mostly lacked visible hemozoin. Finally, while most of the FBS-cultured parasites did not progress beyond lung and early liver stage, HS-cultured parasites reached sexually dimorphic stages by week 6, albeit at a slightly delayed rate compared to in vivo development. In the mouse model, parasites become dimorphic by day 21 post-infection (∼3 weeks)12. Taken together, this protocol allows early sexual development of schistosomes to be studied in vitro, provides a method for high throughput drug screening targeting parasite development, and reduces the reliance on mammalian hosts in research.

Results

Sexually dimorphic schistosomes developed entirely in vitro from cercariae

Searching for culture media able to cultivate parasites from cercariae to maturity, we decided to test Human Serum (HS) compared to the commonly used Foetal Bovine Serum (FBS). This rationale was based both on the fact that humans are the definitive host of Schistosoma mansoni, and that two previous reports in 1981 showed that HS was capable of producing mature schistosomes in vitro32,33, but they were never fully replicated. More recently, culture media were complemented with HS in studies focused on early stages of parasite development for drug testing37–39; however, these worms never progressed beyond the liver stage43. Therefore, to ascertain the in vitro sexual dimorphism establishment of schistosomes entirely developed from cercariae, we compared the effect of two different sources of blood serum: FBS and HS. The development of schistosomula derived from mechanically transformed cercariae was assessed in at least 15 independent experiments, five of which were maintained over a period of at least 10 weeks to assess parasite survival and ability to mate and produce fertile eggs (Figure 1A; Supplementary Table S1).

Dimorphic female and male schistosomes entirely developed in vitro from cercariae.

A. Schematic representation of the collection and mechanical transformation of cercariae into schistosomula for long-term in vitro culture. B. Morphological scoring of cultured S. mansoni schistosomula at the indicated time points after in vitro transformation (week 1 to week 10) for worms in Long Term Culture (LTC) medium supplemented either with Human Serum (HS - left) or Foetal Bovine Serum (FBS - right). Heatmap columns represent five distinct morphological categories, and rows indicate 5 independent culture experiments, i.e. parasites obtained from different batches of infected snails. Heatmap colours represent the percentage of worms for each replicate in each morphological category. Middle panel: Representative images of in vitro schistosomula cultured in either HS or FBS as indicated. Scale bars: 100 µm. Category 1: Early schistosomula; Category 2: Lung schistosomula; Category 3: Early liver schistosomula; Category 4: Late liver schistosomula; Category 5: Dimorphic schistosomula. A detailed description of the developmental categories and representative images are provided in Supplementary Figure S1.

No morphological differences were observed between parasites cultured either in FBS or HS within the first week in culture; in both conditions most parasites were classified as early schistosomula [category 1: 76% ± 30 (average ± SD) in FBS and 73% ± 29 (average ± SD) in HS] with few lung (category 2) and early liver schistosomula (category 3) (Figure 1B, week 1; Supplementary Figure S1). The mean mortality (category 0) at week 1 was slightly higher, but not statistically significant (P= 0.42), in worms cultured in HS [9.75% ± 2.76 (average ± SD)] compared to the mortality registered in FBS-cultured parasites [5.52% ± 5.18 (average ± SD), Supplementary Table S6], consistent with previous findings39.

Differences in parasite development between the two conditions became apparent by week 2 (Figure 1B). At this time point, 14.8% ± 24.9 (average ± SD, excluding dead worms) or 36% ± 33.6 (average ± SD, excluding dead worms) of the parasites cultured in FBS or HS, respectively, have reached category 3, i.e., early liver schistosomulum. Parasites in FBS rarely progressed beyond this stage during the 10-week experiment, with very few parasites (<0.1% ± 0.2, average ± SD) reaching category 4, i.e., late liver schistosomulum. In contrast, worms cultured in HS developed over time across all categories, achieving marked sexual dimorphism by week 6 (13.4% ± 18.6, average ± SD) (Figure 1B; Supplementary Figure S3A), as confirmed by PCR (Supplementary Figure S3B; Supplementary Table S2). No differences in the timing for sexual dimorphism establishment were observed between male and female parasites. The mortality rate of FBS-cultured parasites reached an average of 76.24% ± 23.46 (average ± SD) by week 10, after which the experiments under this condition were stopped as most parasites were dead (Supplementary Figure S2). From that time point onwards only parasites in HS were kept in culture. As previously described for the in vivo development of schistosomes12, in vitro cultured parasites showed developmental asynchrony in agreement with Basch’s observations33; however, by week 10 most of the worms in HS (73.7% ± 25.4, average ± SD) acquired an evident sexual dimorphism (Figure 1B).

The differences in parasite development between the two tested conditions were also confirmed by measuring worm areas at different time points in culture (Figure 2; Supplementary Table S3). Parasites cultured in the presence of HS grew exponentially over time to reach on average 110,069 μm2 ± 72,959 (average ± SD), ∼20-fold larger than parasites cultured in FBS for 10 weeks, which only grew slightly, plateauing at an average 5,417 μm2 ± 1,322 (average ± SD), similar to the size reached by parasites after only two weeks in culture with HS (P-value < 0.01, Supplementary Table S6; Figures 1 and 2).

Parasites cultured in Human Serum grew in size unlike those cultured in Foetal Bovine Serum.

Bar plot representing area measurements of schistosomula developed in vitro and cultured in media complemented either in Foetal Bovine Serum (FBS) or Human Serum (HS) at indicated weeks after cercarial transformation. Bars indicate mean area (µm²) ± SEM. HS: Human Serum (light brown); FBS: Foetal Bovine Serum (blue). *: P-value < 0.01 in indicated pairwise comparisons (Supplementary Table S6).

Parasites developed in human serum readily digest red blood cells

It is widely accepted that within the mammalian host, schistosomes begin to feed on blood cells through their digestive tract at 10 days post-infection; at this time the majority of the parasites have left the lungs and started to colonise the portal system veins in the liver44. Based on both previous reports45, and pilot experiments in which adding human Red Blood Cells (hRBCs) to the culture before day ∼10 did not show obvious haemoglobin digestion, we decided to supplement the culture media with hRBCs at day 13. The addition of hRBCs allowed the parasites to feed and thus continue their development19. At this point, they began to swallow and degrade erythrocytes, producing hemozoin, a black pigment derived from host haemoglobin degradation and visible in the worms’ intestines. Even though few parasites in FBS reached the early liver stage (category 3) within the first week in culture, a minority of them were able to digest hRBCs (3.6% ± 4.7, average ± SD), indicated by displaying black guts (BG) (Figure 3). In contrast, a significant proportion of HS-cultured parasites (36.2% ± 33.6, average ± SD) had already reached the early liver stage by the second week in 3 out of 5 experiments (Figure 1B, week 2). Moreover, parasites cultured in HS displayed a functional digestive system capable of assimilating hRBCs; more than half of the HS-cultured parasites (65% ± 6, average ± SD, P-value < 0.05) showed BG in comparison to FBS-cultured parasites (3.7 ± 4.7 %) (Figure 3; Supplementary Table S4).

Parasites developed in Human Serum readily digest red blood cells.

A. Bar Plot representing the percentage of Human Serum (HS)- or Foetal Bovine Serum (FBS)-cultured schistosomula with (blue bar) or without (light brown bar) black guts (BG) due to the presence of intestinal hemozoin. Washed human Red Blood Cells (hRBCs) were added into the media at day 13 post-transformation and images captured one or two days later. Error Bars = SEM. (Statistical analyses in Supplementary Table S6.) B. Representative images of in vitro developed schistosomula cultured in FBS or HS one day after adding hRBC (+ RBC) and controls without RBC (- RBC). Scale bars: 100 µm.

Development of parasites in human serum may be driven by stem cell proliferation

The growth and development of an organism is driven by a finely regulated combination of cell hyperplasia, hypertrophy, and apoptosis across different tissues46. In schistosomes, a complex stem cell system consisting of both somatic and germline stem cells has been described by leveraging recent single cell transcriptomic data across different developmental stages, including schistosomula and adult worms47. Therefore, we decided to investigate whether the differences in development, growth and feeding capacity between parasites cultured in either FBS- or HS- complemented medium were associated with distinct cellular proliferation rates. EdU pulse experiments revealed notably lower cell proliferation in FBS-cultured parasites as early as 2 days post transformation compared to HS-cultured parasites (Figure 4). It has previously been demonstrated that a group of 5 well-defined somatic stem cells is the only set of cells that actively proliferate in 2-day schistosomula48. Hence, the proliferating cells observed and quantified in our study were most likely stem cells (Supplementary Figure S4; Supplementary Video S1). The difference between the number of proliferating stem cells in parasites cultured in FBS or HS increased significantly over time. HS-cultured schistosomula showed higher numbers of proliferating stem cells, with a median of >48 and >60 EdU+ cells per worm at days 8 and 15, respectively (Figure 4). On the other hand, most FBS-cultured parasites displayed no more than an average of 20 EdU+ cells per worm (Figure 4). hRBCs were added at day 13 post-cercarial transformation to both FBS- and HS- complemented culture media. Worms kept in FBS or HS in the absence of hRBCs were included as controls. Regardless of the serum employed, no significant differences in the numbers of proliferating cells were observed between worms cultured in the presence or absence of hRBCs (Supplementary Tables S5 and S6).

Development of parasites in human serum may be driven by stem cell proliferation.

A. Violin plots showing the number of EdU+ cells per worm at indicated time points (2, 8, and 15 days post cercarial transformation) in parasites cultured either in Foetal Bovine Serum (FBS, blue) or Human Serum (HS, light brown). Human Red Blood Cells (hRBCs) were added in the culture at day 13 post cercarial transformation. The small black dots indicate individual worms, and the big black point indicates the median of EdU+ cells per worm. All worms showing Ill 60 EdU+ cells were counted and clustered together in the group named ‘60 EdU+ cells’. Hence, the data were treated as ordinal, and statistical analysis performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 (*) considered significant (Supplementary Tables S5 and S6). B. Representative images of parasites displaying Edu+ cells at each indicated time point and culture condition. Edu+ cells and nuclei were labelled with Alexa fluor 488 (green) and DAPI (white/grey), respectively. Scale bars: 50 µm or 75 µm as indicated.

In vitro cultured schistosomes display sexual dimorphism, developing reproductive systems and pairing capacity

In vitro-developed schistosomes began to show sexually dimorphic features from day ∼42 onwards in HS-supplemented culture medium (Figure 1, week 6). Confocal microscopy of DAPI- and Phalloidin- stained in vitro cultured male worms confirmed the presence of the gynaecophoric canal, developing 3 to 5 testis lobes with sperm cells, and cirrus (Figure 5A-D; Supplementary Video S2). Likewise, we confirmed the presence of primordial ovaries, oviduct, ootype and uterus in female parasites entirely developed in vitro (Figure 5E-H; Supplementary Video S3). In some experiments sexually dimorphic parasites cultured in HS medium were kept alive for more than 150 days (Supplementary Figure S5). While the establishment of sexual dimorphism was robust and reproducible in more than 15 independent experiments, pairing between male and female parasites was rare. Pairing was observed only in experiments lasting more than 80 days in which we were only able to observe a few couples pairing up, and in addition, this pairing was temporary (Figures 6A, 6B; Supplementary Video S4).

In vitro cultured schistosomes display sexual dimorphism and developing reproductive systems.

Representative confocal microscopy images of in vitro developed male (A-D) and female (E-H) at day 60 in culture. Worms shown in panels G and H are different individuals. CellMask Green Actin Tracking Stain: green, DAPI: grey (A-E, G), cyan (F, H). C: Cirrus. T: Testis, OS: Oral sucker. VS: Ventral sucker. GC: Gynaecophoric canal. G: Gut. GP: Genital pore. U: Uterus. O: Ovary. Oo: Ootype. OV: Oviduc. S: sperm. Scale bar: 50 µm.

In vitro cultured schistosomes are capable of pairing

A, B. Representative bright-field images of pairs of schistosomes developed entirely in vitro after 80 (A) and 150 (B) days of culture in culture medium supplemented with HS. Scale bars: 500 µm (A), 100 µm (B). C, D. Representative bright-field images of a worm pairs between ex vivo-collected males and in vitro-developed females (C) and ex vivo-collected females and in vitro-developed males (D) within 24 hours after placing the worms in the same well to facilitate pairing. Scale bars: 100 µm, E, F. Confocal microscopy image of a schistosome pair (ex vivo-collected male and in vitro-developed female) in copula (E), and magnification of the ovarian area, highlighting maturing oocytes (F). CellMask Green Actin Tracking Stain: green, DAPI: grey, CellMask Deep Red Plasma Membrane Stain: magenta. Scale bar: 150 µm (E), 50 µm (F).

Considering the rarity of the pairing between in vitro developed male and female worms, we investigated whether these parasites display the capacity of pairing with in vivo developed worms collected from experimentally infected mice. Male and female adult worms were recovered from mice by portal perfusion on day 42 post-infection, sorted by sex and placed in culture with worms of the opposite sex developed in vitro. Within 24 hours of initiating the co-culturing of in vitro developed worms with ex vivo collected worms, couples were observed (Figure 6C, D; Supplementary Video S5). These findings suggested that the in vitro developed and sexually dimorphic parasites were capable of intersexual pairing. Moreover, in vitro developed females coupled with ex vivo collected mature males displayed signs of primordial ovary maturation with larger oocytes towards the posterior region of the ovary (Figure 6E, F; Supplementary Video S6). On the other hand, females developed in vitro but not paired with ex vivo males remained immature. Remarkably, in more than 30 independent in vitro culture experiments, where male and female parasites developed sexual dimorphism and eventually paired up, no eggs were produced or laid. This indicates that further refinements in the culture protocol are needed to advance parasite development (the in vitro development was delayed compared to the in vivo development), increase the likelihood of pairing, and facilitate the production of eggs.

Discussion

Recent progress in culture systems and functional genomic tools is allowing researchers to address long standing questions in helminth biology that were previously inaccessible to experimentation49–53. Among these enduring questions, stand those related to the unique sexual biology of the trematode family Schistosomatidae. Schistosomes are dioecious with genetically determined female and male individuals, which is unusual for flatworms5,54. However, the sexual dimorphism of male and female schistosome worms only becomes established within the mammalian host, and represents a critical step towards worm maturation, intersexual pairing, and egg production55. A better understanding of the molecular and cellular basis underlying sexual dimorphism establishment in schistosomes would lead to approaches to block parasite development and life cycle propagation. Studying the sexual development of schistosomes by performing controlled experiments in which parasites can be co-cultured, genetically manipulated or treated with different compounds requires robust and reproducible protocols for long-term in vitro culture. Culture protocols have recently been developed to maintain ex vivo schistosomes collected from infected mice35,36, or to obtain juvenile parasites from cercariae for drug screening experiments37–39,56. However, no long-term culture conditions have been successfully refined to experimentally assess sexual differentiation and dimorphism establishment in schistosomes. Here, aiming to refine a culture medium formulation that supports in vitro parasite development and establishment of sexual dimorphism, we compared the effect of modified Basch’s media19, supplemented with human Red Blood Cells (hRBCs) and 20% of either Fetal Bovine Serum (FBS) or Human Serum (HS). While initial parasite survival and development appeared comparable in both conditions during the first week in culture, striking morphological differences emerged from week 2 onwards. Parasites cultured in HS, progressed through all developmental categories, and acquired sexual dimorphism by week 6. On the other hand, parasites maintained in FBS were stunted at early stages (mainly lung stage). These experimental outcomes were consistent with the findings reported by Paul F. Basch32 in 1980. Probably, from the 1980s onwards a combination of factors that include ethical concerns, variability associated with human-derived products and safety considerations related to blood borne viruses such as HIV, and Hepatitis B and C, determined that parasitologists favoured the use of FBS over HS for helminth culture. FBS has been extensively used in cell culture media, refined and adapted to diverse human and animal cell types since 195857 providing a reliable source of amino acids, carbohydrates, hormones, lipids, proteins, vitamins, and growth factors58. Nevertheless, the scientific community is increasingly advocating for the replacement of FBS as a supplement in tissue culture59. This trend is driven by limitations associated with FBS, including the presence of undefined tentatively harmful factors for the cells in culture, lack of reproducibility, lack of transparency in its production, and critically ethical concerns related to animal welfare60. Our protocol, that supports the fully in vitro development of schistosomes, would also positively impact the 3Rs (i.e., Reduction, Replacement, Refinement)42 for minimising the use of animals for research and serum production61,62.

Parasites cultured in the presence of HS not only developed into sexually dimorphic male and female worms, but strikingly, were able to digest hRBCs and process haemoglobin within one day after the addition of the cells. These findings may reflect differences in gastrodermis and oesophageal gland development; worms cultured in the presence of HS displayed a differentiated gastrodermis that enabled haemoglobin digestion and accumulation of hemozoin within the intestines63. In contrast, most parasites cultured in FBS were unable to produce hemozoin in the presence of hRBCs. This may be due to impaired gastrodermis differentiation that ultimately would lead to stunted development and death. Hematophagous parasites including Plasmodium species and schistosomes, obtain key nutrients via the proteolysis of host hemoglobin. However, this process leads to the production of free-heme groups which in turn generate highly toxic oxygen free radicals and lipid peroxidation64. These toxic free-heme derivatives become inactive when aggregated into an inert crystalline polymer, named hemozoin, observed as dark pigment within the intestines of schistosome intra-mammalian stages63. In addition, increasing evidence shows that hemozoin may play critical roles during parasite development by supplying iron for egg production63, and interaction with the mammalian host by an immune modulatory role65. In the mouse model, schistosomula begin to feed on blood once they have left the lungs and reached the portal system 9 to 11 days post infection44,66. Moreover, schistosomula collected from experimentally infected mice at day 13 post infection already show hemozoin in their guts12. Similarly, HS-developed schistosomula were able to feed on blood in vitro within a comparable time window; hRBCs were added in the medium on day 13 in culture, and within 24 hours most of the parasites’ gut contained hemozoin. These findings suggest that our in vitro culture system successfully recapitulates in vivo gastrodermis development of schistosomula. A better understanding of the in vitro development of the parasite oesophageal gland and gastrodermis and its role in the production of hemozoin will expose tentative novel targets for control67.

Although previous reports have shown that it is possible to cultivate parasites in cell-free environments39,43, adding hRBC has proven essential for long-term maintenance of parasites and for establishing sexual dimorphism. EdU pulse experiments suggested that schistosome in vitro development in the presence of HS may be driven by the proliferation and differentiation of stem cells. The stem cell system in schistosomes has been extensively studied since the first description of somatic stem cells in adult worms68, followed by single cell transcriptomic identification and functional characterisation of three key stem cell populations in intra-snail stages48. Two of these stem cell types, maintained in intra-mammalian stages, proliferate and differentiate into precursors of somatic and germ line cells throughout development14. Although we have not performed transcriptomic analyses in this study, the positive correlation between number of EdU+ cells, and growth (determined by increasing worms’ area) and gross development in HS-developed parasites was evident already within the first 48 hours in culture. These findings indicate that stem cells may play a central role in driving organised growth, tissue differentiation and ultimately the establishment of the sexual dimorphism in HS-developed parasites. Conversely, FBS-cultured parasites displayed no more than 20 EdU+ cells per worm in average, reaching a plateau in the number of proliferating cells from day 8 in culture, which probably underlie the arrested development of these worms. Overall, the FBS-cultured parasites did not progress beyond the lung stage. Consistent with these findings, confocal microscopy imaging revealed in HS-developed parasites clear somatic and reproductive anatomical structures, including male gynaecophoric canals and testes with sperm cells, as well as female ovary primordia, oviducts, ootypes and uteri. These developmental features illustrate the capacity of our culture system to capture key biological transitions previously only accessible from experimentally infected animal models. That said, while our system was highly efficient in producing sexually dimorphic worms, spontaneous pairing between male and female parasites was extremely rare, mainly in aged in vitro cultures (from 80 to 100 days in culture) indicating that other factors, e.g., cholesterol, may be missing35. In any case, we decided to test the pairing capacity of the dimorphic worms by conducting pairing experiments between in vitro-developed females and ex vivo-collected males, and vice versa. Strikingly, couples were observed within 24 hours after co-cultivating in vivo and ex vivo opposite sex parasites, albeit no eggs were produced. However, we observed evidence of pairing-trigger oocyte maturation in in vitro-developed females paired with ex vivo-collected male worms. Our research group has recently been involved in the discovery and functional characterisation of a transcription factor of the retinoic acid receptor family, SmRAR, and related genes69. SmRAR and associated genes may play a key role in oocyte differentiation triggered by pairing. Our pairing experiments and confocal imaging analyses suggested that in vitro developed female oocytes started to differentiate after pairing with an in vivo developed male worm, probably mediated by the activation of the SmRAR pathway69. Moreover, the involvement of a male-derived nonribosomal peptide pheromone, i.e., β-alanyl-tryptamine or ‘BATT’, in the pairing-driven female sexual maturation was demonstrated6. In addition, soluble factors produced by the worms with effect on the opposite sex cannot be ruled out7. It has been suggested that host-derived factors, such as the cytokine transforming growth factor β (TGF-β) may be critical for female reproductive development and embryogenesis70. More recently, other host-derived factors that include cholesterol and ascorbic acid have been shown to be critical for maintaining ex vivo fully mature females collected from infected mice35. Considering these elements in future experiments will help overcome the limitations encountered in this study, including the low rate of spontaneous pairing between in vitro–developed male and female worms and the requirement for extended culture periods (>70 days). In addition, further research is needed to assess the role of host- and parasite-derived cues in schistosome development71.

In summary, we have demonstrated that the presence of HS is essential to support fully in vitro development of S. mansoni parasites from mechanically transformed cercariae to sexually dimorphic adults. Differences in the numbers of proliferating stem cells between worms developed in HS versus FBS were observed as early as 48 hours in culture. These findings may raise some concerns about studies that rely on in vitro culture protocols using FBS, including those focused on parasite developmental biology, interaction with the host, and drug screening. Human serum contains essential host-specific molecules that are absent or insufficient in FBS, and future efforts should aim to identify these factors. Optimising schistosome long-term culture systems will: (1) deepen our understanding of fundamental aspects of schistosome biology, development and host-parasite molecular crosstalk; (2) positively impact the 3Rs principles for animal research; and (3) reveal tentative targets for novel control strategies.

Material and Methods

Ethics statement

The complete life cycle of Schistosoma mansoni (NMRI strain) was maintained at the Wellcome Sanger Institute (WSI) and Aberystwyth University (AU), by infecting Biomphalaria glabrata snails and outbred mice (TO strain). All the animal regulated procedures were conducted under Home Office Project Licences P77E8A062 (WSI) and PP2955700 (AU), following the ARRIVE guidelines (https://arriveguidelines.org) and in accordance with guidelines and regulations stated by the UK Animals (Scientific Procedures) Act 1986 Amendment Regulations 2012. All the protocols were presented and approved by the Animal Welfare and Ethical Review Bodies (AWERB) of the WSI and AU. The AWERB is constituted as required by the UK Animals (Scientific Procedures) Act 1986 Amendment Regulations 2012. Human blood products, including serum and red blood cells were obtained from NHS Blood and Transplant (NHSBT), Non-Clinical Issue (NCI) services for research purposes. NHSBT-NCI provides donated material surplus to clinical requirements or unsuitable for therapeutic use that has been appropriately consented. The supply chain complies with all statutory and regulatory obligations including (but not limited to) the Human Tissue Act (2004) and associated Codes of Practice. The NHSBT-NCI products were acquired under the Study ‘In vitro development of the human parasite Schistosoma’, Integrated Research Application System (IRAS) project ID 319400, protocol number AU/DLS/011, Research Ethics Committee (REC) reference number 23/SS/0017.

Cercariae transformation and culture of schistosomula

Schistosoma mansoni (NMRI strain) schistosomula were obtained and cultured as previously described with minor modifications19. In brief, patent mix strain of Biomphalaria glabrata snails72 experimentally infected en masse with ∼20 miracidia per snail, were thoroughly rinsed, transferred to Lepple water (∼100ml) and exposed to light for 2 h at 26 °C to induce cercarial shedding. Cercariae were collected, passed through a 100μm filter into 50 ml tubes to remove snail debris and faeces, placed 1 h on ice, and concentrated by centrifugation (300 g for 3 min at 4°C with half break deceleration). Cercariae were washed 3 times in 1X PBS supplemented with 200 U/ml penicillin, 200 μg/ml streptomycin, 500 ng/ml amphotericin B (Merck). During the washes, the cercarial pellets were successively combined and finally resuspended in ‘Schistosomula wash medium’ (Dulbecco’s modified Eagle medium - DMEM, supplemented with 10 mM Hepes, 200 U/ml penicillin, 200 g/ml streptomycin, 500 ng/ml amphotericin B). Cercariae were vortexed full speed for 30 sec and placed on ice for 1 min. Thereafter, the cercarial tails were sheared off by ∼10 back and forth passes through a 19-gauge (19-G) needle, a ∼3 μl aliquot of parasites inspected under microscope to confirm that >90% of cercariae had their tail removed, and the cercarial heads were separated from the tails by a Percoll (Merck) gradient (1.5:1 percoll:DMEM) and centrifugation (300g for 15 min at 4 °C, half break deceleration). The pellet containing purified cercarial heads, was collected and washed 3 times by centrifugation (300g for 3 min at 4°C) in ‘Schistosomula wash medium’. Newly transformed schistosomula were counted in 12 5μl-aliquots under the microscope. To reduce parasite mortality due to a high density of worms, ∼8,000 schistosomula were transferred to each well of a 6-well tissue culture plate (Fisher Scientific, Loughborough, UK) containing 5 ml of modified Basch’s medium18 supplemented with additives and FBS or HS heat-inactivated FBS or HS serum, i.e., ‘Long-term culture medium’ (LTC) as indicated in Table 1. The number of parasites cultured per well (∼8,000 schistosomula) was determined empirically, as no formal titration experiments were performed. At higher densities (>10,000 per well), more frequent media changes were required, and parasite development appeared to be impaired. Parasites were cultured in a tissue culture incubator at 37°C under 5% CO2 in air. LTC medium was replaced twice a week and washed human red blood cells (hRBCs) added to a final concentration of 0.02% v/v at day 13 after transformation. Washed hRBCs were replaced every two weeks, or sooner if their numbers decreased due to consumption.

All human blood products, obtained from NHSBT-NCI, were handled with universal precautions within a biosafety cabinet class II and suitable PPE. The HS was heat-inactivated at 56°C for 30 min in an incubator, aliquoted and stored at −20°C until use. The hRBC were washed in ‘Schistosomula wash medium’ and stored at 4°C until use as previously described19. Briefly, the hRBCs were transferred from the pack to 50 ml tubes and washed by centrifugation at 500g for 5 min at 4°C. Half the supernatant was removed, tubes’ content combined, centrifuged as above and resuspended in ∼25 ml of ‘Schistosomula wash medium’. The washes were repeated 4 more times.

Phenotype scoring of developing schistosomula

To analyse the in vitro development of schistosomula, images and videos were taken at regular intervals of at least once a week in >10 independent culture experiments, using a digital Euromex camera connected to an Olympus CK2 optical microscope. Up to 18 pictures per experiment with an average of ∼400 worms per picture were used for parasite staging from at least 5 independent experiments (Supplementary Table S1). The investigators who assigned categories were blinded to experimental conditions. Based on a well-defined staging system12,33, the parasites were classified into 6 developmental categories as described39 (Supplementary Figure S1); category 0: dead parasites showing granulation, rounded shape and degraded or broken tegument - the number of dead parasites may be underestimated due to the loss of some dead worms during the media change; category 1: newly transformed schistosomula, corresponding to the in vivo ‘skin stage’, these worms retain the cercarial head shape with no visible internal structures such as gut; category 2: lung stage schistosomula, elongated and slim worms, usually with a bulged end; category 3: early liver stage, bigger and wider worms compared to the previous stage, with hemozoin pigment (i.e., haemoglobin degradation product) and two ceca gut; category 4: late liver stage, in which the gut is further developed, the two intestinal ceca have fused behind the ventral sucker, and the parasites started to acquire an evident vermiform shape; category 5: sexually dimorphic stage, comprising large, vermiform male- and female- looking worms. The females are longer and thinner than males, the ceca fusion localises closer to the anterior end of the worm (i.e., ∼⅓ of the worm length from the anterior end), and both suckers are smaller compared to the male suckers. The male worms are characterised by a larger and wider head, larger suckers, the ceca fusion closer to the posterior end of the animal (i.e., ∼⅔ of the worm length from the anterior end), and a clearly visible gynecophoral groove12,33. To evaluate differences in mortality between HS- and FBS-cultured parasites, data from 5 experiments were combined and analysed using a Shapiro-Wilk normality test to test normality of the data and a non-parametric Wilcoxon rank sum exact test (Supplementary Tables S1 and S6).

To quantify positive or negative parasites for the presence of hemozoin within their intestines, i.e., black guts + (BG+) or BG- parasites, respectively, hRBCs were added to the culture at day 13, and one or two days later bright field microscopy images were taken. Counts of BG+ and BG- parasites were collected from 5 independent in vitro culture experiments (Supplementary Table S4). Statistical analysis was performed using a Wilcoxon rank-sum (Mann-Whitney) test, with P≤0.05 considered significant (Supplementary Table S6).

Measurement of parasite area

Images of in vitro developing parasites were taken at weeks 1, 2, 3, 4, 6, 8 and 10 after cercariae transformation with an Euromex camera fitted to an Olympus CK2 optical microscope at 4x and 10x magnification. For these magnifications images of scale bars were captured for pixel to area values (µm²). Parasites from 5 independent long-term culture experiments were masked in COCO format from captured images using the CVAT (Computer Vision Annotation Tool) online server (https://www.cvat.ai/). Area metrics were then calculated for each parasite by leveraging the pycocotools package in a customised Python script. Raw and processed data is provided in Supplementary Table S3. Normal distribution was checked using Shapiro test and Mann Whitney tests performed with a significant value set up for P≤0.05 (Supplementary Table S6).

PCR for sexing and confirmation of sexual dimorphism

To confirm the in vitro establishment of sexual dimorphism, parasites were individually collected from culture by day ∼80, for blind sex assignment by bright field microscopy and confirmation by sex-specific PCR73 (Supplementary Figure S3B). Briefly, pictures from individual in vitro developed parasites were taken and sex assigned based on morphological features. Thereafter, genomic DNA from the sex-assigned individual parasites was extracted by lysing the worms at 95°C for 1h in 50mM NaOH, 0.4mM disodium ethylenediaminetetraacetic acid (EDTA), followed by the addition of neutralising buffer (80 mM Tris–HCl in water). Samples were stored at 4°C until use. PCR reactions were performed in a final volume of 10μl consisting of 5μl Platinum 2X hot start PCR master mix (Invitrogen, California, USA), 0.5μl (10μM stock) of forward and reverse primers targeting W1 repeat74 and actin75 (Supplementary Table S2), 1μl of template DNA and 3μl of water. The PCR program comprised an initial denature step at 94°C for 2 min followed by 30 cycles of denature at 98°C for 5 secs, and annealing/elongation at 60°C for 15 secs. The PCR products were resolved on a 1% agarose gel using 1% TAE buffer and 1kb GeneRuler DNA ladder (Thermo Scientific, UK).

Detection and quantification of proliferating cells

Parasites cultured for 15 days (D15) in media supplemented with either HS or FBS and hRBC added at D13 were harvested at days 2, 8 or 15 after cercariae transformation. Parasites cultured under the same conditions but with no hRBC were included as controls. The day before each indicated time point, a solution of 5-ethynyl-2’-deoxyuridine (EdU) was added to the culture medium at 10 μM final concentration. At each indicated time point, parasites from each group were collected and transferred to a 35 µm mesh basket (CEM Microwave Tech) in a well of a 24-well plate, washed in 1x PBS, fixed in 4% Paraformaldehyde (PFA) in 1x PBS for 30 min, washed in 1x PBS and stored at 4°C until use. The worms were incubated in PBST (1x PBS + 0.3% Triton x-100) containing 2μl of Proteinase K (20 mg/ml) for 5 min at room temperature, washed three times in PBST, fixed in 4% PFA in 1x PBS for 10 min at room temperature and washed 5 times in PBST. EdU-positive cells were revealed by incubating the parasites for 30 min in the dark in a solution containing Azide fluor 488 (Sigma-Aldrich), diluted in a solution of 1mM CuSo4, 0.025mM Azide fluor 488, 95mM L-Ascorbic Acid in 1x PBS. The worms were washed 5 times in PBST and incubated in mounting media containing DAPI (Fluoromount-G™ Mounting Medium, with DAPI, Invitrogen) before mounting for confocal microscopy (Leica SP8 super resolution laser confocal microscope). EdU+ cells per parasite were counted for an average of 100 parasites across three independent experiments (Supplementary Table S4). Worms were grouped based on the number of cells per individual, but all those showing Ill60 EdU+ cells were counted in the same group named ‘60 EdU+ cells’. Therefore, the data were considered ordinal and the statistical analysis performed by Kruskal-Wallis test with Dunn multiple comparison post-hoc test, with P≤0.05 considered significant (Supplementary Table S6).

Pairing experiments

Schistosoma mansoni adult male and female worms were collected from experimentally infected mice at 47 days post infection19. Briefly, mice were euthanised by intraperitoneal injection of 200 μl of 200 mg/ml pentobarbital supplemented with 100 U/ml heparin, and worms collected by portal perfusion (the hepatic portal vein was sectioned followed by intracardiac perfusion with phenol-red-free DMEM, containing 10 U/mL heparin) and washed in DMEM. In 24-well plates, in vitro developed female worms were cultured in the presence of in vivo developed male worms and vice versa at a ratio of 1 female every 2 males in the well containing culture medium (Table 1) supplemented with 20% HS. The parasites were cultured for several days in an incubator at 37°C, 5 % CO2 and checked daily under microscope for the presence of pairing.

Parasite staining and confocal imaging

Live parasites collected from culture at indicated time points were stained with 1:1000 dilutions of CellMaskTM Green Actin Tracking Stain (Invitrogen) and CellMaskTM Deep Red Plasma Membrane Stain (Invitrogen) to label polymerized/filamentous actin (F-actin) and cell membranes, respectively. Nuclei were stained by adding two drops per millilitre of NucBlueTM Live ReadyProbesTM Reagent (Invitrogen). After an overnight incubation at 37°C, worms were collected in 15 ml tubes, washed by gravity 3 times in 1x PBS, fixed in 4% PFA (in 1x PBS) overnight at 4°C, washed 3 times in 1x PBS and stored in 200 μl of Fluoromount-G, with DAPI (Invitrogen) before mounting on slides for confocal microscopy. All the images were acquired using a Leica SP8 super resolution laser confocal microscope.